---
title: "MGD56_FJB"
author: "Erika Carlson"
format: html
editor_options: 
  chunk_output_type: console
---

------

This document contains research notes and code generated by:

> **Erika R. Carlson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> erika.carlson@utexas.edu

------



# Overview

This document contains custom functions for analyses. 
Run these code chunks prior to using other documents.


# Preliminaries

- Load {tidyverse}

```{r}
#| include: false
#| message: false
library(tidyverse) # for data wrangling
library(glue) # for data wrangling
library(rstatix) # for data analysis
library(afex) # for data analysis

# Indicate path to save results
results_path <- "data/results/"

# Indicate path to lab files
lab_path <- "/Users/erc2866/Library/CloudStorage/Box-Box/Nixon Lab/Lab Member Data Files/Erika/MGD/MGD56/FJB/"
```



# Custom functions

## Function to add significance to {afex} ANOVA results

`sig_sym()` is a function that adds a significance symbol column to an ANOVA 
table and saves the results as a tibble.

It has the following arguments: 

- **aov_res** is a model object of class list containing results of `aov_ez()`
- **path** indicates the destination path to save files. 
  - It can be a named vector of class character or a quoted path e.g., "data/results/"
  
The function does the following:

- Creates **name_aov_res**, a string, from input **aov_res**, a model object list.
  - This will be used to name the output file
- Format the ANOVA results into a table
  - Compute ANOVA tables for the input model object **aov_res**
  - Convert row names containing the model factors to a column with name Factor
  - Convert data frame to a tibble
  - Rename **Pr(>F)** column containing p-values to **p**
  - Add new column **p<.05** with significance symbols for values of **p**
  - Save results as ".csv". 
    - Path indicated by **path** and **name_aov_res**
    
Dependencies: {tidyverse}

```{r}
aov_sig_sym <- function(aov_res, path) {
  
  # Create object of aov_res in quotes to name file
  name_aov_res <- deparse(substitute(aov_res))
  
  # Format ANOVA table, add significance symbol column, save
  res <- anova(aov_res) %>%
  rownames_to_column(var = "Factor") %>%
  as_tibble() %>% 
  rename(p = "Pr(>F)") %>% 
  mutate("p<.05" = case_when(p < 0.0001 ~ "****",
                             p < 0.001 ~ "***",
                             p < 0.01 ~ "**",
                             p < 0.05 ~ "*",
                             p >= 0.05 ~ "ns")) %>% 
  write_csv(., str_c(path, name_aov_res, ".csv"))
  
  return(res)
}
```


## Function to add significance to {afex} emmeans results

`emm_sig_sym()` is a function that adds a significance symbol column to an 
emmeans contrasts tibble and saves the results as a ".csv".

It has the following arguments: 

- **emm_res** is a tibble containing contrasts from `emmeans()` model object
- **path** indicates the destination path to save files. 
  - It can be a named vector of class character or a quoted path e.g., "data/results/"
  
The function does the following:

- Creates **name_emm_res**, a string, from input **emm_res**, 
a tibble containing contrasts from `emmeans()` model object.
  - This will be used to name the output file
- Format the emmeans results into a table
  - Convert data frame to a tibble (redundant)
  - Add new column **p<.05** with significance symbols for values of **p**
  - Save results as ".csv". 
    - Path indicated by **path** and **name_emm_res**
    
Dependencies: {tidyverse}

```{r}
emm_sig_sym <- function(emm_res, path) {
  
  # Create object of emm_res in quotes to name file
  name_emm_res <- deparse(substitute(emm_res))
  
  # Format emmeans table, add significance symbol column, save
  res <- emm_res %>%
  as_tibble() %>%
  mutate("p<.05" = case_when(p.adj < 0.0001 ~ "****",
                             p.adj < 0.001 ~ "***",
                             p.adj < 0.01 ~ "**",
                             p.adj < 0.05 ~ "*",
                             p.adj >= 0.05 ~ "ns")) %>% 
  write_csv(., str_c(path, name_emm_res, ".csv"))
  
  return(res)
}
```


## Function to extract ANOVA results sentence from {afex} anova model object

`add_anova_summary()` is a function to extract anova sentence and append to *anova_summary_df*

It has the following arguments: 

- **aov_res** is a model object of class list containing results of `aov_ez()`
- **data_name** is the quoted name of the data set to analyze (e.g., "measure_region")

Dependencies: {dplyr}, {glue}, {afex}

```{r}
# Dependencies
# library(dplyr)
# library(glue)
# library(afex)

# Function to extract sentence and append to **anova_summary_df**
add_anova_summary <- function(aov_res, data_name) {
  
  # Extract and format ANOVA sentence
  anova_tbl <- anova(aov_res) %>%
    rownames_to_column(var = "Factor") %>%
    as_tibble()

  anova_sentence <- anova_tbl %>%
    mutate(
      F_val = sprintf("%.2f", `F`),
      ges_val = sprintf("%.3f", `ges`),
      p_val = sprintf("%.4f", `Pr(>F)`),
      num.Df = sprintf("%.0f", `num Df`),
      den.Df = sprintf("%.0f", `den Df`)
      ) %>%
    transmute(
      text = glue("{Factor} (F{num.Df},{den.Df} = {F_val}, \u03B72G = {ges_val}, p = {p_val})")
      ) %>%
    pull(text) %>%
    paste(collapse = ", ") %>%
    paste0("ANOVA yielded a main effect of ", .)

  # Capture name of anova object
  aov_name <- deparse(substitute(aov_res))
  
   # Create new row
  new_row <- tibble(
    data_frame = data_name,
    anova_object = aov_name,
    anova_sentence = anova_sentence
  )

  # Create or append to summary df
  if (!exists("anova_summary_df", envir = .GlobalEnv)) {
    assign("anova_summary_df", new_row, envir = .GlobalEnv)
  } else {
    anova_summary_df <<- bind_rows(anova_summary_df, new_row)
  }

  # Return current state
  return(anova_summary_df)
}
```

```{r}
#| eval: false

# Example usage
aov3 <- aov_ez(id = 'subject', 
               dv = "hippo", 
               data = df, 
               between = c('diet', 'drug', 'sex'))


anova_summary_df <- add_anova_summary(aov_res = aov3,
                                      data_name = "fjb_hippo")
```
