---
title: "MGD56_MicrogliaMorphology"
author: "Erika Carlson"
format: html
editor_options: 
  chunk_output_type: console
---

------

This document contains research notes and code generated by:

> **Erika R. Carlson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> erika.carlson@utexas.edu

------


# Overview

We examined the role of microglia in alcohol-induced neurodegeneration in a rat model of an AUD by depleting microglia with PLX5622 (PLX), a highly selective antagonist of colony-stimulating factor-1 receptor (CSF1R), a receptor kinase critical for survival and proliferation of microglia/macrophages.

Iba1 (ionized calcium binding adapter 1) is a marker of microglia/macrophages.
FJB (FluoroJade-B) is a marker of dying neurons.

Adult female and male sprague-dawley rats were injected with PLX [50 mg/kg, 10 mL/kg, subcutaneous (s.c.), every 12 h] or vehicle for 11 days. During the last 4 days of PLX, rats received ethanol (0-5 g/kg, 25% EtOH w/v, intragastric (i.g.), every 8 h) in diluted Vanilla Ensure Plus (50% w/v) in a Majchrowicz paradigm. Tissue was collected during intoxication, 3-6 h post-EtOH (T0).
Blood ethanol concentration (BEC; mg/dL) was determined from tail blood collected 90 min post-dose 7.
Tissue was prepared by phosphate-buffered saline perfusion (PBS), brain extraction and bisection, at least 24 h post-fixation in 4% paraformaldehyde (PFA) at 4°C, and storage in PBS at 4°C until sliced 1:12 series at 40 um thickness on vibrating microtome. Tissue slices were stored in cryoprotectant at -20°C until immunohistochemical assay.

Slides were blinded prior to imaging and unblinded prior to MicrogliaMorphology ImageJ analysis due to file naming conventions.

- Blinding and unblinding code found in "Slide_blinding.qmd"

Microglia were analyzed using the MicrogliaMorphology ImageJ macros and {MicrogliaMorphologyR} package

- Preprocessing (output in "SingleChannelTiffs_Input/") using "MicrogliaMorphology_Preprocessing_Macro_2025-12-18.ijm" macro

```
run("8-bit"); //convert from RBG to 8-bit image
run("Invert"); //invert image for dark background
run("Set Scale...", "distance=144 known=50.06 unit=µm");
```

MicrogliaMorphology detailed here:

- https://github.com/ciernialab/MicrogliaMorphology/tree/main
- https://www.youtube.com/watch?v=YhLCdlFLzk8

For BioVoxel_Toolbox > Threshold Check > Checked items:

- bright objects on dark background
- ignore black (default=off)
- Quantification (relative)

Batch1 (dark), Batch2 (light)

- Sensitivity: 82
- Hunang does not do well with few microglia
- Triangle is a bit sparse but does separate
- Copied "SingleCells/" folder to repository "data/results/MicrogliaMorphology/batch<#>/" to read files into FracLac (issue with cloud-based storage?)

".ijm" files required some edits to work which usually involved splitting nested arguments into two lines (detailed in "/MicrogliaMorphology/README.docx"). Used individual MM steps. 



# Preliminaries

References 

- https://github.com/ciernialab/MicrogliaMorphologyR
- https://ciernialab.github.io/MicrogliaMorphologyR/articles/MicrogliaMorphologyR.html

Install {MicrogliaMorphologyR}
```{r}
# install devtools package and make sure you have the latest version
install.packages("devtools")

# install MicrogliaMorphologyR using devtools
devtools::install_github('ciernialab/MicrogliaMorphologyR')
```

Load packages and set seed for reproducibility of results
```{r}
library(MicrogliaMorphologyR)
library(factoextra)
library(ppclust)
```

```{r}
set.seed(10)
```



# Data Exploration


## Load fraclac and skeleton data, tidy, and merge into final data frame

> *NOTE:* FracLac directory has a subfolder

```{r}
results_path <- "/Users/erc2866/Development/Repos/carlson-2026-plx-ethanol/data/results/MicrogliaMorphology/"

mm.dir <- "/Users/erc2866/Library/CloudStorage/Box-Box/Nixon Lab/Lab Member Data Files/Erika/MGD/MGD56/MicrogliaMorphology/all/"
fraclac.dir <- str_c(mm.dir, "FracLac/20260413013210")
skeleton.dir <- str_c(mm.dir, "SkeletonResults")

# May be time-intensive if analyzing on the order of 1000s of cells
fraclac <- fraclac_tidying(fraclac.dir) 
skeleton <- skeleton_tidying(skeleton.dir)

data <- merge_data(fraclac, skeleton)
# Final data set parameters                                                            
finaldata <- metadata_columns(data, c("Stain","Study","RatID","Sex","Drug","Diet","BrainRegion","SlideSlice","Magnification"), sep="_")
```


## Generate heatmap of correlations across features

Explore the morphology features measured by MicrogliaMorphology and how they relate to each other by generating a heatmap of spearman’s correlations across the 27 different morphology features.
```{r}
featurecorrelations(finaldata,
                    featurestart=12, featureend=38, # start/end of measure cols
                    rthresh=0.8, pthresh=0.05,
                    title="Correlations across features")

# to get the underlying stats depicted in the heatmap
# ensure start/end of measure cols matches finaldata
correlationstats <- featurecorrelations_stats(finaldata,
                                              featurestart=12, featureend=38, 
                                              rthresh=0.8, pthresh=0.05)
correlationstats %>% head()
```

## Exploratory data visualization

```{r}
# gather your numerical morphology data into one column ('measure') which contains the feature name, and another column ('value') which contains measured values
finaldata_gathered <- finaldata %>% gather(measure, value, 12:ncol(finaldata))

# check for outliers
outliers_boxplots(finaldata_gathered)

outliers_distributions(finaldata_gathered)

# checking different normalization features
normalize_logplots(finaldata_gathered,1)

normalize_minmax(finaldata_gathered)

normalize_scaled(finaldata_gathered)

# transform data for downstream analyses
# using logtransformed data as PCA input
finaldata_logtransformed <- transform_log(finaldata, 1, start=12, end=38) 
finaldata_minmaxtransformed <- transform_minmax(finaldata, start=12, end=38)
finaldata_scaled <- transform_scale(finaldata, start=12, end=38)

# get sample size of data based on factors of interest
# "RatID","Sex","Drug","Diet","BrainRegion"
samplesize(finaldata, RatID, Sex, Drug, Diet, BrainRegion)
samplesize(finaldata, Sex, Drug, Diet, BrainRegion)
```


### Obtain microglia number and density per image

```{r}
# list out all variables of interest present in the original image names to obtain numbers of microglia per image
microglianumbers <- samplesize(finaldata, Stain, Study, RatID, Sex, Drug, Diet, BrainRegion, SlideSlice, Magnification) 
microglianumbers %>% print(n=3, width=Inf)

# add Name column back in -- make sure that the resulting strings in the Name column match the names of your original input .tiff files that you used for MicrogliaMorphology !!
microglianumbers <- microglianumbers %>% unite("Name", Stain:Magnification, sep="_", remove=FALSE) 
microglianumbers %>% print(n=3, width=Inf)

# path to Areas.csv file
AreasPath <- str_c(mm.dir, "Areas.csv")

# use celldensity function to calculate density at image-level: values are under the "Density" column
Density <- celldensity(AreasPath, microglianumbers)
#> Warning: Expected 2 pieces. Additional pieces discarded in 148 rows [1, 2, 3, 4, 5, 6,
#> 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, ...].
Density %>% print(n=5, width=Inf)

# Group by variables of interest
Density %>% 
  group_by(RatID, Sex, Drug, Diet, BrainRegion) %>%
  summarise(num=sum(num), Area=sum(Area)) %>% # calculate new cell numbers and new areas at the brain region level
  mutate(Density=num/Area) # calculate new density at the brain region level

write_csv(Density, str_c(results_path, "iba1_density.csv"))
```

## PCA

```{r}
pcadata_elbow(finaldata_logtransformed, featurestart=12, featureend=38)

pca_data <- pcadata(finaldata_logtransformed, featurestart=12, featureend=38,
                    pc.start=1, pc.end=10)
head(pca_data,3)
```

### Generate heatmap of correlations between PCs and features

```{r}
pcfeaturecorrelations(pca_data, pc.start=1, pc.end=3, 
                      feature.start=22, feature.end=48, 
                      rthresh=0.75, pthresh=0.05, 
                      title="Correlation between PCs and features")

# to get the underlying stats depicted in the heatmap above
correlationstats <- pcfeaturecorrelations_stats(pca_data, pc.start=1, pc.end=3,
                                                feature.start=22, feature.end=48,
                                                rthresh=0.75, pthresh=0.05)
correlationstats %>% head()
```

### Visually explore sources of variability

```{r}
# gather your data by experimental variables (e.g., Treatment, Sex, MouseID, etc.)
gathered_expvariables <- pca_data %>% gather(variable, value, 13:17) 

plots_expvariable(gathered_expvariables, "PC1", "PC2")
```

### K-means clustering on PCs

```{r}
# Prepare the data

## for k-means clustering: scale PCs 1-3, which together describe ~85% of variability
pca_data_scale <- transform_scale(pca_data, start=1, end=3) # scale pca data as input for k-means clustering
kmeans_input <- pca_data_scale[1:3]



# Cluster optimization prior to running k-means

# check for optimal number of clusters using wss and silhouette methods
sampling <- kmeans_input[sample(nrow(kmeans_input), 1000),] # sample 5000 random rows for cluster optimization

fviz_nbclust(sampling, kmeans, method = 'wss', nstart=25, iter.max=50) # 4 clusters

fviz_nbclust(sampling, kmeans, method = 'silhouette', nstart=25, iter.max=50) # 4 clusters
```


#### Fuzzy clustering (skip)

Computationally intensive

```{r, eval = FALSE}
# cluster and combine with original data
data_kmeans <- fcm(kmeans_input, centers=4, nstart=25)
pca_kmeans <- cbind(pca_data[1:5], data_kmeans)
```

#### Regular k-means (hard clustering)

```{r}
# cluster and combine with original data
data_kmeans <- kmeans(kmeans_input, centers=4)

# Here, we are creating a new data frame that contains the first 2 PCs and original dataset, then renaming the data_kmeans$cluster column to simply say "Cluster". You can bind together as many of the PCs as you want. Binding the original, untransformed data is useful if you want to plot the raw values of any individual morphology measures downstream. 
pca_kmeans <- cbind(pca_data[1:2], finaldata, as.data.frame(data_kmeans$cluster)) %>%
  rename(Cluster=`data_kmeans$cluster`) 
head(pca_kmeans,3)
```

### Plot k-means clusters in PC space

```{r}
plot <- clusterplots(pca_kmeans, "PC1", "PC2")
plot

# customization example
plot + scale_colour_viridis_d() # customizeable example: add color scheme of choice 
```

### Cluster-specific measures on average for each morphology feature, relative to other clusters

```{r}
clusterfeatures(pca_kmeans, featurestart=14, featureend=40)
```

Compare the individual features across clusters to characterize the clusters:

- Cluster 1 = ramified (largest territory span and branching complexity)
- Cluster 2 = hypertrophic (average territory span, high branch thickness --> via pixel density in hull)
- Cluster 3 = rod-like (greatest oblongness, lowest circularity)
- Cluster 4 = ameboid (lowest territory span, high circularity, smallest branch lengths)




# ColorByCluster

### Formatting data for ColorByCluster input (Color coding in ImageJ)

One image at a time

```{r}
# isolate out all the cells for your specific image of interest
colorbycluster <- pca_kmeans %>% 
  filter(RatID=="05", BrainRegion=="DGML", SlideSlice=="1-10") %>% select(c(Cluster, ID))
head(colorbycluster)

# save ".csv" to feed into ColorByCluster function in MicrogliaMorphology ImageJ macro
write.csv(colorbycluster, str_c(mm.dir, "Rat05_DGML_1-10_data.csv"))
```

### Cluster characterization

```{r}
# calculate cluster percentages across variables of interest
cp <- clusterpercentage(pca_kmeans, "Cluster", RatID, Drug, Diet, Sex, BrainRegion)
cp$Drug <- factor(cp$Drug, levels=c("VEH","PLX"))
cp$Diet <- factor(cp$Diet, levels=c("CON","EtOH"))
cp$Sex <- factor(cp$Sex, levels=c("M","F"))

# update cluster labels
cp <- cp %>% mutate(Cluster = 
                      case_when(Cluster=="1" ~ "Ramified",
                                Cluster=="2" ~ "Hypertrophic",
                                Cluster=="3" ~ "Rod-like",
                                Cluster=="4" ~ "Ameboid"))

# Quick check of cluster proportions when considering experimental variables of interest
cp %>% 
  filter(BrainRegion=="DGML") %>% # in this example, we filter for our brain region of interest
  clusterpercentage_boxplots(Drug, Diet, Sex) # grouping variables

# example graph of data given variables of interest
cp %>% 
  filter(BrainRegion=="DGML") %>%
  ggplot(aes(x=Cluster, y=percentage, group=interaction(Cluster, Drug))) +
  facet_wrap(~Diet) +
  geom_boxplot(aes(group=interaction(Cluster, Drug), fill=Drug)) +
  scale_fill_manual(values=c("#fde725","#482878")) +
  geom_point(position=position_dodge(width=0.8), size=0.75, aes(group=interaction(Cluster,Drug), color=Sex)) +
  ggtitle("PLX binge dataset: K-means clusters") +
  labs(fill="Drug") +
  theme_bw(base_size=14) +
  theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1))
```

## Statistical analysis

### Microglia number and Density

See "MGD56_IBA1.qmd" (note: need to reset working directory)

`"/Users/erc2866/Library/CloudStorage/Box-Box/Nixon Lab/Lab Member Data Files/Erika/MGD/MGD56/MicrogliaMorphology/all/SkeletonResults"`

### Cluster percentage changes at animal level, in response to experimental variables

```{r}
# prepare percentages dataset for downstream analysis
stats.input <- cp
stats.input$RatID <- factor(stats.input$RatID)
stats.input$Cluster <- factor(stats.input$Cluster)
stats.input$Drug <- factor(stats.input$Drug)
stats.input$Diet <- factor(stats.input$Diet)
stats.input$Sex <- factor(stats.input$Sex)

stats.input %>% 
  filter(BrainRegion == "DGML") %>%
  ungroup() %>%
  count(Cluster, Diet, Drug) %>% 
  tidyr::complete(Cluster, Diet, Drug) %>% # check whether any combinations are missing
  print(n = Inf)


# run stats analysis for changes in cluster percentages, at the animal level
# you can specify up to two posthoc comparisons (posthoc1 and posthoc2 arguments) - if you only have one set of posthocs to run, specify the same comparison twice for both arguments. you will just get the same results in output[[2]] and output[[3]].
stats.testing <- stats_cluster.animal(data = stats.input %>% filter(BrainRegion=="DGML") %>% droplevels(), # drop unused factor levels 
                                      model = "percentage ~ Cluster*Diet*Drug + (1|RatID)", 
                                      posthoc1 = "~Drug|Cluster|Diet", 
                                      posthoc2 = "~Drug|Cluster", adjust = "bonferroni")

# reduce model complexity
stats.testing <- stats_cluster.animal(data = stats.input %>% filter(BrainRegion=="DGML"), 
                                      model = "percentage ~ Cluster*Diet + Cluster*Drug + (1|RatID)", 
                                      posthoc1 = "~Drug|Cluster|Diet", 
                                      posthoc2 = "~Drug|Cluster", adjust = "bonferroni")

stats.testing[[1]] # anova
stats.testing[[2]] # posthoc 1
stats.testing[[3]] # posthoc 2
stats.testing[[4]] # DHARMa model check
stats.testing[[5]] # summary of model
```

### Individual morphology measures, at the animal level (averaged across cells for each animal)

```{r}

```



## Reset working directory

```{r}
setwd("~/Development/Repos/carlson-2026-plx-ethanol")
```



# Archive

You can add options to executable code like this 

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
