---
title: "MGD5and6_Analyses"
author: "Erika Carlson"
format: html
editor_options: 
  chunk_output_type: console
---

------

This document contains research notes and code generated by:

> **Erika R. Carlson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> erika.carlson@utexas.edu

With assistance from:

> **K. Ryan Thompson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> kt25399@utexas.edu

------

# Overview

We examined the role of microglia in alcohol-induced neurodegeneration in a rat model of an AUD by depleting microglia with PLX5622 (PLX), a highly selective antagonist of colony-stimulating factor-1 receptor (CSF1R), a receptor kinase critical for survival and proliferation of microglia/macrophages.

Iba1 (ionized calcium binding adapter 1) is a marker of microglia/macrophages.
FJB (FluoroJade-B) is a marker of dying neurons.

Adult female and male sprague-dawley rats were injected with PLX [50 mg/kg, 10 mL/kg, subcutaneous (s.c.), every 12 h] or vehicle for 11 days. During the last 4 days of PLX, rats received ethanol (0-5 g/kg, 25% EtOH w/v, intragastric (i.g.), every 8 h) in diluted Vanilla Ensure Plus (50% w/v) in a Majchrowicz paradigm. Tissue was collected during intoxication, 3-6 h post-EtOH (T0).
Blood ethanol concentration (BEC; mg/dL) was determined from tail blood collected 90 min post-dose 7.
Tissue was prepared by phosphate-buffered saline perfusion (PBS), brain extraction and bisection, and at least 24 h post-fix in 4% paraformaldehyde (PFA) at 4°C before storing in PBS at 4°C until slicing at 40 um thickness on vibrating microtome.


# Preliminaries

- Load packages
- Indicate path to save results

```{r}
#| include: false
#| message: false
library(tidyverse)
library(rstatix) # for statistical analysis
library(emmeans) # for statistical analysis
library(lme4) # for statistical analysis
library(afex) # dependency for ANOVA analysis
library(ggplot2) # for data visualization
library(ggpubr) # for data visualization

# Indicate path to save results
results_path <- "data/results/"
```

## Function to add significance to {afex} ANOVA results

`sig_sym()` is a function that adds a significance symbol column to an ANOVA table and saves the results as a tibble.

It has the following arguments: 

- `aov_res` is a model object of class list containing results of `aov_ez()`

It uses the following external objects:

- **results_path** indicates the destination path to save files. "data/results/"
  - Could modify function to include this as an argument
  
The function does the following:

- Creates **name_aov_res**, a string, from input `aov_res`, a model object list.
  - This will be used to name the output file
- Format the ANOVA results into a table
  - Compute ANOVA tables for the input model object `aov_res`
  - Convert row names containing the model factors to a column with name Factor
  - Convert data frame to a tibble
  - Rename **Pr(>F)** column containing p-values to **p**
  - Add new column **p<.05** with significance symbols for values of **p**
  - Save results as ".csv". 
    - Path indicated by **results_path** external object and **name_aov_res**

```{r}
aov_sig_sym <- function(aov_res) {
  
  # Create object of aov_res in quotes to name file
  name_aov_res <- deparse(substitute(aov_res))
  
  # Format ANOVA table, add significance symbol column, save
  anova(aov_res) %>%
  rownames_to_column(var = "Factor") %>%
  as_tibble() %>% 
  rename(p = "Pr(>F)") %>% 
  mutate("p<.05" = case_when(p < 0.0001 ~ "****",
                             p < 0.001 ~ "***",
                             p < 0.01 ~ "**",
                             p < 0.05 ~ "*",
                             p >= 0.05 ~ "ns")) %>% 
  write_csv(., str_c(results_path, name_aov_res, ".csv"))
}
```


```{r}
emm_sig_sym <- function(emm_res) {
  
  # Create object of emm_res in quotes to name file
  name_emm_res <- deparse(substitute(emm_res))
  
  # Format emmeans table, add significance symbol column, save
  emm_res %>%
  as_tibble() %>%
  mutate("p<.05" = case_when(p.value < 0.0001 ~ "****",
                             p.value < 0.001 ~ "***",
                             p.value < 0.01 ~ "**",
                             p.value < 0.05 ~ "*",
                             p.value >= 0.05 ~ "ns")) %>% 
  write_csv(., str_c(results_path, name_emm_res, ".csv"))
}
```

# Binge statistics

**2025-09-15**
Currently excluding rats that died during binge: 17, 36, 31, 28
May later exclude #35 (hind limb/paw edema) <- would need to update exclude column in all data files

```{r Load and process body weights}
# Load weight data
weight <- read_csv("data/weight.csv") %>% filter(exclude == "N")

# Check data
slice_sample(weight, n = 6)

# Get summary statistics per day per sex
weight_common <- weight %>% 
  group_by(sex) %>% 
  get_summary_stats(day_01:day_11, type = "common") %>% 
  as_tibble() %>% 
  write_csv(., str_c(results_path, "weight_summary.csv"))

# Obtain mean ± se onset weight overall
weight_onset <- weight %>% 
  group_by(sex) %>% 
  get_summary_stats(day_01, type = "mean_se") %>% 
  as_tibble() %>% 
  write_csv(., str_c(results_path, "weight_onset.csv"))


# Compute two-way ANOVA on average onset weight in males using {afex} package
weight_onset_aov2_m <- aov_ez(id = 'subject', 
                              dv = 'day_01', 
                              data = weight %>% filter(sex == 'M'), 
                              between = c('diet', 'drug')
                              )
weight_onset_aov2_m

# Add significance symbol and save result as tibble using custom `sig_sym()` function
aov_sig_sym(weight_onset_aov2_m)
# no significance


# Repeat in females
weight_onset_aov2_f <- aov_ez(id = 'subject', 
                              dv = 'day_01', 
                              data = weight %>% filter(sex == 'F'), 
                              between = c('diet', 'drug')
                              )
weight_onset_aov2_f

aov_sig_sym(weight_onset_aov2_f)
# no significance


# Create weight_long table for analysis across days
weight_long <- weight %>% pivot_longer(cols = day_01:day_11, names_to = "day", values_to = "weight")


# Compute three-way ANOVA on average weight over time in males using {afex} package
weight_aov3_m <- aov_ez(id = 'subject', 
                        dv = 'weight', 
                        data = weight_long %>% filter(sex == 'M'), 
                        between = c('diet', 'drug'),
                        within = 'day'
                              )
weight_aov3_m

# Use custom function `sig_sym() to add significance symbol and save result as tibble
aov_sig_sym(weight_aov3_m)
# main effect of drug, day
# interaction of drug x day, diet x day, drug x diet x day


# Repeat in females
weight_aov3_f <- aov_ez(id = 'subject', 
                        dv = 'weight', 
                        data = weight_long %>% filter(sex == 'F'), 
                        between = c('diet', 'drug'),
                        within = 'day'
                              )
weight_aov3_f

aov_sig_sym(weight_aov3_f)
# main effect of day
# interaction of drug x day, diet x day


###############################################################################


# Load % baseline weight data
percent_weight <- read_csv("data/percent_baseline_weight.csv") %>% filter(exclude == "N")

# Check data
slice_sample(percent_weight, n = 6)

# Create percent_weight_long table for analysis across days
percent_weight_long <- percent_weight %>% pivot_longer(cols = day_01:day_11, names_to = "day", values_to = "weight")


# Compute three-way ANOVA on average weight over time in males using {afex} package
percent_weight_aov3_m <- aov_ez(id = 'subject', 
                        dv = 'weight', 
                        data = percent_weight_long %>% filter(sex == 'M'), 
                        between = c('diet', 'drug'),
                        within = 'day'
                              )
percent_weight_aov3_m

# Use custom function `sig_sym()` to add significance symbol and save result as tibble
aov_sig_sym(percent_weight_aov3_m)
# main effect of drug, day
# interaction of drug x day, diet x day, drug x diet x day

# Repeat in females
percent_weight_aov3_f <- aov_ez(id = 'subject', 
                        dv = 'weight', 
                        data = percent_weight_long %>% filter(sex == 'F'), 
                        between = c('diet', 'drug'),
                        within = 'day'
                              )
percent_weight_aov3_f

aov_sig_sym(percent_weight_aov3_f)
# main effect of drug, day
# interaction of drug x day, diet x day, drug x diet x day
```


```{r Load and process intoxication scores}
# Load behavioral intoxication score data
intox <- read_csv("data/intoxication.csv") %>% filter(exclude == "N")

# Check data
slice_sample(intox, n = 6)

# Calculate average behavioral intoxication score per subject
intox <- intox %>%
  mutate(mean = rowMeans(select(., score_01:score_12)))

# Get summary statistics
intox_summary <- intox %>%
  group_by(drug, sex) %>%
  get_summary_stats(mean, type = "mean_se") %>% 
  as_tibble() %>% 
  write_csv(., str_c(results_path, "/intoxication_summary.csv"))
intox_summary


# Compute two-way ANOVA on average behavioral intoxication using {afex} package
intox_aov2 <- aov_ez(id = 'subject', 
                     dv = 'mean', 
                     data = intox, 
                     between = c('sex', 'drug')
                     )
intox_aov2
# Main effect of sex, drug
# No interaction

# Use custom function `sig_sym()` to add significance label and save result as tibble
aov_sig_sym(intox_aov2)

# Reformat data for three-way ANOVA across dose sessions (time)
intox_long <- intox %>% select(!mean) %>% 
  pivot_longer(cols = score_01:score_12, 
               names_to = "session",
               values_to = "score")

# Compute three-way ANOVA on behavioral intoxication across time using {afex} package
intox_aov3 <- aov_ez(id = 'subject', 
                     dv = 'score', 
                     data = intox_long, 
                     between = c('sex', 'drug'),
                     within = 'session'
                     )
intox_aov3
# Main effect of sex, drug, session
# sex x session, drug x session, sex x drug x session interaction

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(intox_aov3)


# Post-hoc comparisons, could restrict to common factors (no interaction) with $contrasts[c()]
# within session: [1] F PLX - M PLX [2] F PLX - F VEH [3] F PLX - M VEH [4] M PLX - F VEH [5] M PLX - M VEH [6] F PLX - M PLX
# 72 comparisons total
# Note: score_01 currently included but yields NA (all rats score 0)
intox_aov3_post <- emmeans(intox_aov3, 
                           pairwise ~ sex * drug | session,
                           adjust = "bonferroni")$contrasts %>% 
  as_tibble()

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(intox_aov3_post)

# Sig "main effect" diffs: score_04 F PLX - M PLX, score_10 F PLX - M PLX, score_10 F PLX - F VEH, score_11 F PLX - F VEH
# also some interaction diffs: M PLX - F VEH, F PLX - M VEH
```
