---
title: "MGD56_MicrogliaMorphology"
author: "Erika Carlson"
format: html
editor_options: 
  chunk_output_type: console
---

------

This document contains research notes and code generated by:

> **Erika R. Carlson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> erika.carlson@utexas.edu

------


# Overview

We examined the role of microglia in alcohol-induced neurodegeneration in a rat model of an AUD by depleting microglia with PLX5622 (PLX), a highly selective antagonist of colony-stimulating factor-1 receptor (CSF1R), a receptor kinase critical for survival and proliferation of microglia/macrophages.

Iba1 (ionized calcium binding adapter 1) is a marker of microglia/macrophages.
FJB (FluoroJade-B) is a marker of dying neurons.

Adult female and male sprague-dawley rats were injected with PLX [50 mg/kg, 10 mL/kg, subcutaneous (s.c.), every 12 h] or vehicle for 11 days. During the last 4 days of PLX, rats received ethanol (0-5 g/kg, 25% EtOH w/v, intragastric (i.g.), every 8 h) in diluted Vanilla Ensure Plus (50% w/v) in a Majchrowicz paradigm. Tissue was collected during intoxication, 3-6 h post-EtOH (T0).
Blood ethanol concentration (BEC; mg/dL) was determined from tail blood collected 90 min post-dose 7.
Tissue was prepared by phosphate-buffered saline perfusion (PBS), brain extraction and bisection, at least 24 h post-fixation in 4% paraformaldehyde (PFA) at 4°C, and storage in PBS at 4°C until sliced 1:12 series at 40 um thickness on vibrating microtome. Tissue slices were stored in cryoprotectant at -20°C until immunohistochemical assay.

Slides were blinded prior to imaging and unblinded prior to MicrogliaMorphology ImageJ analysis due to file naming conventions.
- Blinding and unblinding code found in Slide_blinding.qmd

Microglia were analyzed using the MicrogliaMorphology ImageJ macros and {MicrogliaMorphologyR} package

MicrogliaMorphology detailed here:

- https://github.com/ciernialab/MicrogliaMorphology/tree/main
- https://www.youtube.com/watch?v=YhLCdlFLzk8

.ijm files required some edits to work which usually involved splitting nested arguments into two lines (detailed in "MicrogliaMorphology_README_ERC.docx"). Individual MM steps worked best. 



# Preliminaries

References 

- https://github.com/ciernialab/MicrogliaMorphologyR
- https://ciernialab.github.io/MicrogliaMorphologyR/articles/MicrogliaMorphologyR.html

Install {MicrogliaMorphologyR}
```{r}
# install devtools package and make sure you have the latest version
install.packages("devtools")

# install MicrogliaMorphologyR using devtools
devtools::install_github('ciernialab/MicrogliaMorphologyR')
```

Load packages and set seed for reproducibility of results
```{r}
library(MicrogliaMorphologyR)
library(factoextra)
library(ppclust)
```

```{r}
set.seed(10)
```



# Data Exploration


## Load fraclac and skeleton data, tidy, and merge into final data frame

> *NOTE:* FracLac directory has a subfolder

```{r}
fraclac.dir <- "/Users/erc2866/Library/CloudStorage/Box-Box/Nixon Lab/Lab Member Data Files/Erika/MGD/MGD56/MicrogliaMorphology/FracLac/20251219030337"
skeleton.dir <- "/Users/erc2866/Library/CloudStorage/Box-Box/Nixon Lab/Lab Member Data Files/Erika/MGD/MGD56/MicrogliaMorphology/SkeletonResults"

# May be time-intensive if analyzing on the order of 1000s of cells
fraclac <- fraclac_tidying(fraclac.dir) 
skeleton <- skeleton_tidying(skeleton.dir)

data <- merge_data(fraclac, skeleton)
finaldata <- metadata_columns(data, c("Stain","Study","RatID-Slide-Slice","BrainRegion","Magnification"), sep="_")

# Final data set parameters                                                            
# finaldata <- metadata_columns(data, c("Stain","Study","RatID","Sex","Drug","Diet","BrainRegion","Slide-Slice","Magnification"), sep="_")
```


## Generate heatmap of correlations across features

Explore the morphology features measured by MicrogliaMorphology and how they relate to each other by generating a heatmap of spearman’s correlations across the 27 different morphology features.
```{r}
featurecorrelations(finaldata,
                    featurestart=8, featureend=34,
                    rthresh=0.8, pthresh=0.05,
                    title="Correlations across features")

# to get the underlying stats depicted in the heatmap
correlationstats <- featurecorrelations_stats(finaldata,
                                              featurestart=8, featureend=34,
                                              rthresh=0.8, pthresh=0.05)
correlationstats %>% head()
```

## Reset working directory

```{r}
setwd("~/Development/Repos/carlson-2026-plx-ethanol")
```



# Archive

You can add options to executable code like this 

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
