---
title: "MGD56_FJB"
author: "Erika Carlson"
format: html
editor_options: 
  chunk_output_type: console
---

------

This document contains research notes and code generated by:

> **Erika R. Carlson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> erika.carlson@utexas.edu

------



# Overview

We examined the role of microglia in alcohol-induced neurodegeneration in a rat model of an AUD by depleting microglia with PLX5622 (PLX), a highly selective antagonist of colony-stimulating factor-1 receptor (CSF1R), a receptor kinase critical for survival and proliferation of microglia/macrophages.

Iba1 (ionized calcium binding adapter 1) is a marker of microglia/macrophages.
FJB (FluoroJade-B) is a marker of dying neurons.

Adult female and male sprague-dawley rats were injected with PLX [50 mg/kg, 10 mL/kg, subcutaneous (s.c.), every 12 h] or vehicle for 11 days. During the last 4 days of PLX, rats received ethanol (0-5 g/kg, 25% EtOH w/v, intragastric (i.g.), every 8 h) in diluted Vanilla Ensure Plus (50% w/v) in a Majchrowicz paradigm. Tissue was collected during intoxication, 3-6 h post-EtOH (T0).
Blood ethanol concentration (BEC; mg/dL) was determined from tail blood collected 90 min post-dose 7.
Tissue was prepared by phosphate-buffered saline perfusion (PBS), brain extraction and bisection, at least 24 h post-fixation in 4% paraformaldehyde (PFA) at 4°C, and storage in PBS at 4°C until sliced 1:12 series at 40 um thickness on vibrating microtome. Tissue slices were stored in cryoprotectant at -20°C until histological assay.

1:12 series of vibratome-cut (40 um) slices were subjected to IHC for:
- Iba1 DAB, imaged at 20X (4X representative)
- FJB, surveyed by hand, representative staining imaged at 40X

FJB slides were blinded for analysis in "Slide_blinding.qmd". The following code: 

- compiles the raw counts into usable format
- unblinds data via blind_key

**Data frame edits:**

- WWW-1-5, vHippo, "R" (a crossed out 1) removed
  - `raw_counts %>% filter(is.na(as.numeric(vHippo)) & !is.na(vHippo))`
- EE-2-4, pirif-apirif, 0 substituted for "?"
  - `raw_counts %>% filter(is.na(as.numeric(pirif_apirif)) & !is.na(pirif_apirif))`


# Preliminaries

- Load {tidyverse}

```{r}
#| include: false
#| message: false
library(tidyverse) # for data wrangling
library(glue) # for data wrangling
library(rstatix) # for data analysis
library(afex) # for data analysis

# Indicate path to save results
results_path <- "data/results/"

# Indicate path to lab files
lab_path <- "/Users/erc2866/Library/CloudStorage/Box-Box/Nixon Lab/Lab Member Data Files/Erika/MGD/MGD56/FJB/"
```

## Make sure custom functions from "MGD56_CustomFunctions.qmd" are loaded

# Data tidying

Create *comp* data table consisting of: 

1. Number of sections per brain region per *blind_code* - integer vs NA
2. Sum of FJB+ cell counts per brain region per *blind_code*
3. Average FJB+ cell count per brain region per *blind_code*
- One *blind_code* per row


Read in the ".csv" file containing the raw counts

```{r}
# set path to data
path <- lab_path

# read in data file, rename column *subject* to *blind_code*
raw_counts <- read_csv(str_c(path, "MGD56_FJB_raw.csv")) %>% 
  rename(blind_code = subject) 

# make a column for hippo counts (combine vHippo and dHippo)
raw_counts <- raw_counts %>% 
  mutate(hippo = case_when(
    !is.na(dHippo) & !is.na(vHippo) ~ dHippo + vHippo,
    !is.na(dHippo) & is.na(vHippo)  ~ dHippo,
    is.na(dHippo)  & !is.na(vHippo) ~ vHippo,
    .default = NA_real_
  ),
  .before = dHippo)

# save
write_csv(raw_counts, str_c(lab_path, "MGD56_FJB_raw.csv"))
```


Write `count_by_group()`, `sum_by_group()`, `mean_by_group()` functions to 
compile raw counts into usable summaries by subject with the following arguments:

- *df*, a data frame 
- *cols*, a vector of *df* column names to be iterated over
- *group_col*, name of column to group data by
- Using an anonymous function `\(x)` within `summarise()`

See https://adv-r.hadley.nz/fp.html, could update with functional / functions in list

```{r}
# Count the number of non-NA observations of each col given group_col
count_by_group <- function(df, cols, group_col) {
  df %>% group_by({{ group_col }}) %>%
    summarise(across(all_of(cols), \(x) sum(!is.na(x)), .names = "n_{.col}"),
              .groups = "drop")
}


# Sum the values of each col given group_col
sum_by_group <- function(df, cols, group_col) {
  df %>% group_by({{ group_col }}) %>%
    summarise(across(all_of(cols), \(x) sum(x, na.rm = TRUE), .names = "sum_{.col}"), 
              .groups = "drop")
}


# Average the values of each col given group_col
mean_by_group <- function(df, cols, group_col) {
  df %>% group_by({{ group_col }}) %>%
    summarise(across(all_of(cols), \(x) mean(x, na.rm = TRUE), .names = "mean_{.col}"), 
              .groups = "drop")
}
```


Create *fjb_regions* vector of column names of interest, then create *count*, *sum*, and *mean* data frames by calling the above functions. Join the three data frames into *comp* data frame and save.

```{r}
fjb_regions <- names(raw_counts %>% select(-c(blind_code, slide, slice_num, bregma, notes, date)))

# c("cing_retro", "motor", "visual", "parietal", "somato", "auditory", 
# ad_insul", "dHippo", "vHippo", "peri_entor", "pirif_apirif", "PLCo", "BLA")

count <- count_by_group(df = raw_counts, cols = fjb_regions, group_col = blind_code)

sum <- sum_by_group(df = raw_counts, cols = fjb_regions, group_col = blind_code)

avg <- mean_by_group(df = raw_counts, cols = fjb_regions, group_col = blind_code)

comp <- count %>% left_join(sum, by = "blind_code") %>% left_join(avg, by = "blind_code")
```



# Unblind FJB counts

```{r}
blind_key <- read_csv("data/fjb_blind_key.csv")

# add meta data to compiled count data
counts_by_subject <- left_join(blind_key, comp) %>% 
  filter(!is.na(n_cing_retro)) %>% arrange(diet, desc(drug), sex)

# save to repository
write_csv(counts_by_subject, "data/fjb_bysubject.csv")

# save to lab files
write_csv(counts_by_subject, str_c(lab_path, "MGD56_FJB_bysubject.csv"))

# save select regions to graph in prism
fjb_prism <- counts_by_subject %>% 
  select(subject, drug, diet, sex, mean_hippo:mean_pirif_apirif) %>% 
  arrange(diet, desc(drug), sex) %>% 
  unite("group", drug, diet, sex, sep = "_") %>%
  mutate(group = fct_inorder(group)) %>% # preserve arrange order
  pivot_longer(-c(subject, group)) %>% 
  pivot_wider(names_from = c(group, subject), values_from = value)

# save to repository
write_csv(fjb_prism, "data/fjb_prism.csv")

# save to lab files
write_csv(fjb_prism, str_c(lab_path, "fjb_prism.csv"))
```


# Analysis

## Summarize data
**Updated 2026-02-11**

Using custom function `counting_by_group()`, same framework as above

```{r}
counts_by_subject <- read_csv("data/fjb_bysubject.csv")
  

# create vector of column names, restrict to data columns with mean values
col_names <- names(counts_by_subject %>% select(starts_with("mean"))) # modify

# turn into a function
counting_by_group <- function(df, cols, group_cols) {
  
  sem <- function(x) {
    sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
  }
  
  df %>%  
    group_by(across(all_of(group_cols))) %>%
    summarise(n = n(),
              across(all_of(cols),
                     list(mean = \(x) mean(x, na.rm = TRUE),
                          sd = \(x) sd(x, na.rm = TRUE),
                          sem = sem),
                     .names = "{col}_{fn}"),
              .groups = "drop")
}


counts_by_group <- counting_by_group(counts_by_subject, col_names, group_cols = c("drug", "diet", "sex"))

write_csv(counts_by_group, "data/results/fjb_bygroup.csv")

# save to lab files
write_csv(counts_by_group, "/Users/erc2866/Library/CloudStorage/Box-Box/Nixon Lab/Lab Member Data Files/Erika/MGD/MGD56/FJB/MGD56_FJB_bygroup.csv")
```

Determine min/max/mean number of brain slices per region
```{r}
counts_by_subject <- read_csv("data/fjb_bysubject.csv")

# turn into a function
slicing_by_group <- function(df, cols, group_cols) {
  
  df %>%  
    group_by(across(all_of(group_cols))) %>%
    summarise(across(all_of(cols),
                     list(mean = \(x) mean(x, na.rm = TRUE),
                          min = \(x) min(x, na.rm = TRUE),
                          max = \(x) max(x, na.rm = TRUE)),
                     .names = "{col}_{fn}"),
              .groups = "drop")
}

# create vector of column names, restrict to data columns with mean values
col_names <- names(counts_by_subject %>% select(starts_with("n"))) # modify

slices_by_group <- slicing_by_group(counts_by_subject, col_names, group_cols = c("drug", "diet", "sex"))

# save to repository
write_csv(slices_by_group, "data/results/fjb_slices_bygroup.csv")

# save to lab files
write_csv(slices_by_group, str_c(lab_path, "MGD56_FJB_slices_bygroup.csv"))
```


## Statistics
**2026-04-11**

```{r}
# create mean count data frame per subject
fjb <- read_csv("data/fjb_bysubject.csv") %>% 
  select(subject, drug, diet, sex, starts_with("mean_")) %>% 
  rename_with(~ str_remove(.x, "^[^_]*_")) %>% 
  mutate(across(c(drug, diet, sex), as.factor)) # define factors

# pull statistics to report
fjb_summary <- fjb %>%
  group_by(drug, diet, sex) %>%
  summarise(across(hippo:pirif_apirif, 
                   list(
                     mean = mean, 
                     sd = sd,
                     sem = \(x) sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))), 
                   .names = "{col}_{fn}"))

# save to repo
fjb_summary %>% as_tibble() %>% 
  write_csv(., str_c(results_path, "fjb_summary.csv"))

# save to lab files
fjb_summary %>% as_tibble() %>% 
  write_csv(., str_c(lab_path, "fjb_summary.csv"))
```


### Inverse Hyberbolic Sine (ArcSinh) Transformation

- ArcSinh transformation is similar to natural log for values greater than 1
- Curve shifted slightly up (greater y-value) and slightly steeper slope
- Behaves linearly between 0 and 1 and can handle 0s
- https://www.statalist.org/forums/forum/general-stata-discussion/general/1553489-inverse-hyperbolic-sinc-transformation

>**NOTE:** **Zero-inflated Poisson** distrubution is a type of modeling well-suited for count data with lots of zeros. It may be a better approach than transformation. See https://stats.oarc.ucla.edu/r/dae/zip/

Using {ggplot2}

```{r}
# asinh and log data
x_asinh <- seq(-50, 50, length.out = 500)
x_log <- seq(0.01, 50, length.out = 500)
x_log10 <- seq(0.01, 50, length.out = 500)

df_plot <- bind_rows(
  tibble(x = x_asinh, y = asinh(x_asinh), fn = "asinh"),
  tibble(x = x_log, y = log(x_log), fn = "log"),
  tibble(x = x_log10, y = log10(x_log10), fn = "log10"))

# Plot
ggplot(df_plot, aes(x = x, y = y, color = fn)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey50") +
  scale_color_manual(values = c("asinh" = "#185FA5", "log" = "green4", "log10" = "#E07B39"),
                     labels = c("asinh", "log", "log10")) +
  labs(x = "argument", y = "transformed", color = NULL) +
  theme_bw() +
  theme(
    legend.position.inside = c(0.15, 0.85),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = "grey80"),
    legend.key.size = unit(0.5, "cm"),
    axis.title.y = element_text(angle = 0, vjust = 0.5))
```

```{r}
# asinh and log data, zoomed in
x_asinh <- seq(-4, 4, length.out = 100)
x_log <- seq(0.01, 4, length.out = 100)
x_log10 <- seq(0.01, 4, length.out = 100)

df_plot <- bind_rows(
  tibble(x = x_asinh, y = asinh(x_asinh), fn = "asinh"),
  tibble(x = x_log, y = log(x_log), fn = "log"),
  tibble(x = x_log10, y = log10(x_log10), fn = "log10"))

# Plot
ggplot(df_plot, aes(x = x, y = y, color = fn)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey50") +
  scale_color_manual(values = c("asinh" = "#185FA5", "log" = "green4", "log10" = "#E07B39"),
                     labels = c("asinh", "log", "log10")) +
  labs(x = "argument", y = "transformed", color = NULL) +
  theme_bw() +
  theme(
    legend.position.inside = c(0.15, 0.85),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = "grey80"),
    legend.key.size = unit(0.5, "cm"),
    axis.title.y = element_text(angle = 0, vjust = 0.5))
```


Three-way ANOVAs using {afex} package

```{r}
results_path <- "data/results/fjb_asinh/"

cols_to_transform <- names(fjb %>% select(cing_retro:BLA))

asinh_fjb <- fjb %>%
  select(subject:BLA) %>% 
  mutate(across(
  .cols = all_of(cols_to_transform),
  .fns = ~ asinh(.), # handles zeros better than log + constant
  .names = "{.col}"))

# save select regions to graph in prism
asinh_fjb_prism <- asinh_fjb %>% 
  select(subject, drug, diet, sex, hippo:pirif_apirif) %>% 
  arrange(diet, desc(drug), sex) %>% 
  unite("group", drug, diet, sex, sep = "_") %>%
  mutate(group = fct_inorder(group)) %>% # preserve arrange order
  pivot_longer(-c(subject, group)) %>% 
  pivot_wider(names_from = c(group, subject), values_from = value)

# save to repository
write_csv(asinh_fjb_prism, "data/asinh_fjb_prism.csv")

# save to lab files
write_csv(fjb_prism, str_c(lab_path, "asinh_fjb_prism.csv"))

# hippo ------------------------------------------------------------------------
fjb_hippo_aov3 <- aov_ez(id = 'subject', 
                     dv = 'hippo', 
                     data = asinh_fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_hippo_aov3, results_path)
# Main effect of drug, diet (not sex)
# drug x diet interaction

add_anova_summary(fjb_hippo_aov3, "asinh_fjb_hippo")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_hippo_aov3_post <- emmeans(fjb_hippo_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_hippo_aov3_post, results_path)



# dHippo -----------------------------------------------------------------------
fjb_dhippo_aov3 <- aov_ez(id = 'subject', 
                     dv = 'dHippo', 
                     data = asinh_fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_dhippo_aov3, results_path)
# No main effects, drug x diet and diet x sex interactions

add_anova_summary(fjb_dhippo_aov3, "asinh_fjb_dHippo")

# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_dhippo_aov3_post <- emmeans(fjb_dhippo_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_dhippo_aov3_post, results_path)



# vHippo -----------------------------------------------------------------------
fjb_vhippo_aov3 <- aov_ez(id = 'subject', 
                     dv = 'vHippo', 
                     data = asinh_fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_vhippo_aov3, results_path)
# Main effect of drug, diet (not sex)
# drug x diet interaction

add_anova_summary(fjb_vhippo_aov3, "asinh_fjb_vHippo")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_vhippo_aov3_post <- emmeans(fjb_vhippo_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_vhippo_aov3_post, results_path)
# tranding difference PLX CON - PLX EtOH F (p = 0.0545)



# peri_entor -------------------------------------------------------------------
fjb_peri_entor_aov3 <- aov_ez(id = 'subject', 
                     dv = 'peri_entor', 
                     data = asinh_fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_peri_entor_aov3, results_path)
# Main effect of drug, diet (not sex)
# drug x diet interaction

add_anova_summary(fjb_peri_entor_aov3, "asinh_fjb_peri_entor")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_peri_entor_aov3_post <- emmeans(fjb_peri_entor_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_peri_entor_aov3_post, results_path)
# sig PLX CON - PLX EtOH F



# pirif_apirif -----------------------------------------------------------------
fjb_pirif_apirif_aov3 <- aov_ez(id = 'subject', 
                     dv = 'pirif_apirif', 
                     data = asinh_fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_pirif_apirif_aov3, results_path)
# Main effect of diet (not drug or sex)
# drug x diet interaction

add_anova_summary(fjb_pirif_apirif_aov3, "asinh_fjb_pirif_apirif")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_pirif_apirif_aov3_post <- emmeans(fjb_pirif_apirif_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_pirif_apirif_aov3_post, results_path)
```

```{r}
flextable(anova_summary_df) %>% save_as_docx(., path = "data/results/anova_summary_df_fjb.docx")
```



**OTHER ANALYSIS**

Three-way ANOVAs using {afex} package

## Raw data

```{r}
results_path <- "data/results/fjb_raw/"

# hippo ------------------------------------------------------------------------
fjb_hippo_aov3 <- aov_ez(id = 'subject', 
                     dv = 'hippo', 
                     data = fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_hippo_aov3, results_path)
# Main effect of drug, diet (not sex)
# drug x diet interaction

add_anova_summary(fjb_hippo_aov3, "fjb_hippo")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_hippo_aov3_post <- emmeans(fjb_hippo_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_hippo_aov3_post, results_path)


# dhippo -----------------------------------------------------------------------
fjb_dhippo_aov3 <- aov_ez(id = 'subject', 
                     dv = 'dHippo', 
                     data = fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_dhippo_aov3, results_path)
# No main effects or interactions

add_anova_summary(fjb_dhippo_aov3, "fjb_dhippo")


# vhippo -----------------------------------------------------------------------
fjb_vhippo_aov3 <- aov_ez(id = 'subject', 
                     dv = 'vHippo', 
                     data = fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_vhippo_aov3, results_path)
# Main effect of drug, diet (not sex)
# drug x diet interaction

add_anova_summary(fjb_vhippo_aov3, "fjb_vhippo")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_vhippo_aov3_post <- emmeans(fjb_vhippo_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_vhippo_aov3_post, results_path)


# peri_entor -------------------------------------------------------------------
fjb_peri_entor_aov3 <- aov_ez(id = 'subject', 
                     dv = 'peri_entor', 
                     data = fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_peri_entor_aov3, results_path)
# Main effect of drug, diet (not sex)
# drug x diet interaction

add_anova_summary(fjb_peri_entor_aov3, "fjb_peri_entor")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_peri_entor_aov3_post <- emmeans(fjb_peri_entor_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_peri_entor_aov3_post, results_path)


# pirif_apirif -----------------------------------------------------------------
fjb_pirif_apirif_aov3 <- aov_ez(id = 'subject', 
                     dv = 'pirif_apirif', 
                     data = fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_pirif_apirif_aov3, results_path)
# Main effect of diet (not drug or sex)
# drug x diet interaction

add_anova_summary(fjb_pirif_apirif_aov3, "fjb_pirif_apirif")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_pirif_apirif_aov3_post <- emmeans(fjb_pirif_apirif_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_pirif_apirif_aov3_post, results_path)
```



## Log Transformation (+ 0.1)

Three-way ANOVAs using {afex} package

```{r}
results_path <- "data/results/fjb_log/"

cols_to_transform <- names(fjb %>% select(cing_retro:BLA))

log_fjb <- fjb %>%
  select(subject:BLA) %>% 
  mutate(across(
  .cols = all_of(cols_to_transform),
  .fns = ~ log(. + 0.1), # + x handles zeros, can warp data
  .names = "{.col}"))

# hippo ------------------------------------------------------------------------
fjb_hippo_aov3 <- aov_ez(id = 'subject', 
                     dv = 'hippo', 
                     data = log_fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_hippo_aov3, results_path)
# Main effect of drug, diet (not sex)
# drug x diet interaction

add_anova_summary(fjb_hippo_aov3, "log_fjb_hippo")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_hippo_aov3_post <- emmeans(fjb_hippo_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_hippo_aov3_post, results_path)



# dhippo -----------------------------------------------------------------------
fjb_dhippo_aov3 <- aov_ez(id = 'subject', 
                     dv = 'dHippo', 
                     data = log_fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_dhippo_aov3, results_path)
# No main effects, drug x diet and diet x sex interactions

add_anova_summary(fjb_dhippo_aov3, "log_fjb_dhippo")



# vhippo -----------------------------------------------------------------------
fjb_vhippo_aov3 <- aov_ez(id = 'subject', 
                     dv = 'vHippo', 
                     data = log_fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_vhippo_aov3, results_path)
# Main effect of drug, diet (not sex)
# drug x diet interaction

add_anova_summary(fjb_vhippo_aov3, "log_fjb_vhippo")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_vhippo_aov3_post <- emmeans(fjb_vhippo_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_vhippo_aov3_post, results_path)



# peri_entor -------------------------------------------------------------------
fjb_peri_entor_aov3 <- aov_ez(id = 'subject', 
                     dv = 'peri_entor', 
                     data = log_fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_peri_entor_aov3, results_path)
# Main effect of drug, diet (not sex)
# drug x diet interaction, drug x diet x sex interaction (!)

add_anova_summary(fjb_peri_entor_aov3, "log_fjb_peri_entor")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_peri_entor_aov3_post <- emmeans(fjb_peri_entor_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_peri_entor_aov3_post, results_path)



# pirif_apirif -----------------------------------------------------------------
fjb_pirif_apirif_aov3 <- aov_ez(id = 'subject', 
                     dv = 'pirif_apirif', 
                     data = log_fjb, 
                     between = c('drug', 'diet', 'sex')
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(fjb_pirif_apirif_aov3, results_path)
# Main effect of diet (not drug or sex)
# drug x diet interaction

add_anova_summary(fjb_pirif_apirif_aov3, "log_fjb_pirif_apirif")


# Post-hoc comparisons, restrict to common factors (no interaction), within sex
# 8 comparisons total
fjb_pirif_apirif_aov3_post <- emmeans(fjb_pirif_apirif_aov3, 
                           pairwise ~ drug * diet | sex,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("PLX CON - VEH CON",
                         "PLX CON - PLX EtOH", 
                         "VEH CON - VEH EtOH",
                         "PLX EtOH - VEH EtOH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(fjb_pirif_apirif_aov3_post, results_path)
```



~~~~~~~~~~~~START SCRATCH~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

```{r}
# testing ANOVA assumptions

ggplot(fjb, aes(x = hippo)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ drug + diet + sex)

fjb <- fjb %>% mutate(hippo_log = log(hippo + 0.1)) # + x handles zeros, can warp data


fjb <- fjb %>% mutate(hippo_sqrt = sqrt(hippo))


fjb <- fjb %>% mutate(hippo_cbrt = sign(hippo) * abs(hippo)^(1/3))


fjb <- fjb %>% mutate(hippo_ihs = asinh(hippo))


fjb %>%
  pivot_longer(cols = c(hippo, hippo_log, hippo_sqrt, hippo_cbrt, hippo_ihs),
               names_to = "transform", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ transform, scales = "free")

model <- aov(hippo ~ drug * diet * sex, data = fjb)
model_log <- aov(hippo_log ~ drug * diet * sex, data = fjb)
model_sqrt <- aov(hippo_sqrt ~ drug * diet * sex, data = fjb)
model_ihs  <- aov(hippo_ihs  ~ drug * diet * sex, data = fjb)
model_cbrt <- aov(hippo_cbrt ~ drug * diet * sex, data = fjb)

shapiro.test(residuals(model))
shapiro.test(residuals(model_log))
shapiro.test(residuals(model_sqrt))
shapiro.test(residuals(model_ihs))
shapiro.test(residuals(model_cbrt))

# Check if it helped
shapiro.test(residuals(model))
hist(residuals(model))
qqnorm(residuals(model)); qqline(residuals(model))

hist(fjb$hippo_sqrt)
hist(fjb$hippo_log)
hist(fjb$hippo)

hist(fjb$hippo_ihs)

# inverse hyperbolic sine transformation improves data structure, but still non-normal due to zero-inflation
```

```{r}
# permutation test
install.packages("coin")
library(coin)

independence_test(hippo ~ drug * diet * sex, 
                  data = fjb,
                  distribution = approximate(nresample = 10000))

install.packages("lmPerm")
library(lmPerm)


model_perm <- aovp(hippo ~ drug * diet * sex, 
                   data = fjb, 
                   perm = "Prob",
                   maxIter = 10000)
summary(model_perm)

install.packages("pak")
pak::pkg_install("rcompanion")
library(rcompanion)

# Diet effect within each drug * sex combination
diet_results <- fjb %>%
  group_by(drug, sex) %>%
  group_modify(~ {
    test <- pairwisePermutationTest(hippo ~ diet, data = .x, method = "none")
    as.data.frame(test)
  })

# Drug effect within each diet * sex combination  
drug_results <- fjb %>%
  group_by(diet, sex) %>%
  group_modify(~ {
    test <- pairwisePermutationTest(hippo ~ drug, data = .x, method = "none")
    as.data.frame(test)
  })

all_results <- bind_rows(diet_results, drug_results)
all_results$p_adj <- p.adjust(all_results$p.value, method = "bonferroni")

print(all_results)
```

~~~~~~~~~~~~END SCRATCH~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~


# Archived code

## Debugging

```{r echo = TRUE, results = 'hide'}
# Code generated from debugging
```

## Failed strategies

```{r echo = TRUE, results = 'hide'}
# Code and/or pipelines that didn't work & why


# Compiling FJB counts
# `raw` is a data table of counts per brain region (across columns) per slice per subject (listed in blind_rand column)

# gives count per value of AD_insul column, including count of NAs
# want collapsed count of observations in AD_insul column, excluding NAs
tidy <- raw |> group_by(blind_rand) |>
  count(AD_insul, blind_rand, name = "AD_insul_n_slices")

# gives desired count for all values of AD_insul, but need to modify to iterate across all columns
# use `across()`
tidy <- aggregate(cbind(Count = AD_insul) ~ blind_rand, raw, length)

```


### Write `add_anova_summary()` function to extract anova sentence and append to `anova_summary_df
**ANOVA MODEL BUILT INTERNALLY**
```{r}
# Dependencies
# library(dplyr)
# library(glue)
# library(afex)

# Function to extract sentence and append to anova_summary_df
add_anova_summary <- function(
    data, data_name, dv, id, between, within = NULL, anova_object_name,
    path) {
  
  # Run ANOVA
  anova_obj <- aov_ez(
    id = id,
    dv = dv,
    data = data,
    between = between,
    within = within
  )

  # Extract and format ANOVA sentence
  anova_tbl <- anova(anova_obj) %>%
    rownames_to_column(var = "Factor") %>%
    as_tibble()

  anova_sentence <- anova_tbl %>%
    mutate(
      F_val = sprintf("%.2f", `F`),
      ges_val = sprintf("%.3f", `ges`),
      p_val = sprintf("%.4f", `Pr(>F)`),
      num.Df = sprintf("%.0f", `num Df`),
      den.Df = sprintf("%.0f", `den Df`)
      ) %>%
    transmute(
      text = glue("{Factor} (F{num.Df},{den.Df} = {F_val}, \u03B72G = {ges_val}, p = {p_val})")
      ) %>%
    pull(text) %>%
    paste(collapse = ", ") %>%
    paste0("ANOVA yielded a main effect of ", .)

  # Create new row
  new_row <- tibble(
    data_frame = data_name,
    anova_object = anova_object_name,
    anova_sentence = anova_sentence
  )

  # Create or append to summary df
  if (!exists("anova_summary_df", envir = .GlobalEnv)) {
    assign("anova_summary_df", new_row, envir = .GlobalEnv)
  } else {
    anova_summary_df <<- bind_rows(anova_summary_df, new_row)
  }

  # Return current state
  return(anova_summary_df)
}
```

```{r eval = FALSE}
# --- Example usage ---

# Load data and clean
df <- read_csv("data/fjb_bysubject.csv") %>% 
  select(subject, drug, diet, sex, starts_with("mean_")) %>% 
  rename_with(~ str_remove(.x, "^[^_]*_")) %>% 
  mutate(across(c(drug, diet, sex), as.factor)) # define factors

# Run this once for each dataset:
anova_summary_df <- add_anova_summary(
  data = df,
  data_name = "fjb_hippo",
  dv = "hippo",
  id = "subject",
  between = c("diet", "drug", "sex"),
  anova_object_name = "aov3_hippo"
)

# Print to verify
print(anova_summary_df)
```



# ANOVA version

```{r}
# ignoring violations of normality and homoscedacity
aov3 <- aov_ez(id = 'subject', 
               dv = region, 
               data = fjb, 
               between = c('diet', 'drug', 'sex'))

anova(aov3) %>%
  rownames_to_column(var = "Factor") %>%
  as_tibble() %>% 
  write_csv(., str_c(results_path, paste("fjb", region, "aov3.csv", sep = "_")))


# post-hoc comparisons via estimated marginal means
aov3_emm <- emmeans(aov3, ~ diet * drug, adjust = "bonferroni")

# simple effects
pairs(aov3_emm, adjust = "bonferroni") %>% as_tibble() %>% 
  write_csv(., str_c(results_path, 
                     paste("fjb", region, "aov3_posthoc_simple.csv", sep = "_")))


# # not using
# # post-hoc comparisons restricted to common factors (3) explicitly 
# emm_diet <- emmeans(aov3, ~ diet | drug * sex)
# emm_drug <- emmeans(aov3, ~ drug | diet * sex)
# 
# # Combine contrast lists and adjust for number of comparisons together
# combined <- rbind(pairs(emm_diet), pairs(emm_drug))
# 
# summary(combined, adjust = "bonferroni") %>% as_tibble() %>%
#   write_csv(., str_c(results_path, 
#                      paste("fjb", region, "aov3_posthoc.csv", sep = "_")))
```
