---
title: "MGD56_ResultsTables"
author: "Erika Carlson"
format: html
editor_options: 
  chunk_output_type: console
---

------

This document contains research notes and code generated by:

> **Erika R. Carlson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> erika.carlson@utexas.edu

With assistance from:

> **K. Ryan Thompson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> kt25399@utexas.edu

------

# Overview

This document contains code to format the output of {afex} ANOVAs into results tables.

>NOTE: Column names for 3-way ANOVA posthoc include p.adj, as correction was done after filtering comparisons.


# Preliminaries

- Load packages
- Indicate path to save results

```{r}
#| include: false
#| message: false
library(tidyverse)
library(rstatix) # for statistical analysis
library(emmeans) # for statistical analysis
library(lme4) # for statistical analysis
library(afex) # dependency for ANOVA analysis
library(ggplot2) # for data visualization
library(ggpubr) # for data visualization

# Indicate path to save results
data_path <- "data/"
results_path <- "data/results/"
```


using {flextable}

```{r}
install.packages("flextable")
library(flextable)

init_flextable_defaults() # reset to default

set_flextable_defaults(
  font.family = "Arial",
  font.size = 10,
  text.align = "left",
  digits = 1,
  fun_theme = arial_theme
  )
```


# Functions for table formatting

**ToDo** 

- set overall table width to 7.5"
  - might need to do during export step
- set condition of replacing 0.0000 with <0.0001 for small p values

Dependencies: {flextable}, {tidyverse}

- theme based on https://github.com/idea-labs/comsldpsy

```{r}
# Dependencies: {flextable}, {tidyverse}

# theme based on https://github.com/idea-labs/comsldpsy
arial_theme <- function (ft)  {
  ft %>% 
    flextable::font(fontname = "Arial", part = "all") %>% 
    flextable::fontsize(size = 12, part = "all") %>% 
    flextable::align(align = "left", part = "body") %>% 
    flextable::align(align = "center", part = "header") %>% 
    flextable::rotate(rotation = "lrtb", align = "top", part = "body") %>% 
    flextable::border_remove() %>% 
    flextable::hline_top(border = officer::fp_border(width = 2), 
                         part = "all") %>% 
    flextable::hline_bottom(border = officer::fp_border(width = 2),
                            part = "all") %>% 
    flextable::hline(i = 1, border = officer::fp_border(width = 1), part = "header") %>%
    flextable::set_table_properties(layout = "autofit")
}

# alternative syntax for formatting ANOVA table with `flextable()`
# colformat_double(j = "num Df", digits = 1)

# 2-way ANOVA
ft_aov2 <- function(x, end_title){
  
  z <- flextable(x) %>% 
  # force specific number of decimal places
    set_formatter(
      `num Df` = function(x) sprintf("%.1f", x),
      `den Df` = function(x) sprintf("%.1f", x),
      MSE = function(x) sprintf("%.2f", x),
      `F` = function(x) sprintf("%.2f", x),
      ges = function(x) sprintf("%.3f", x),
      p = function(x) sprintf("%.4f", x)) %>% 
  # more cosmetics
  # set_header_labels() %>% 
    add_footer_lines("* p<0.05, ** p<0.01, *** p<0.001, **** p<0.0001") %>% 
    color(part = "footer", color = "#666666") %>% 
  # custom title
    set_caption(as_paragraph(
      as_chunk(paste0("Two-way ANOVA of ", end_title),
               props = fp_text_default(font.family = "Arial")))) %>% 
    # custom width for first column
    width(j = 1, width = 1.2)
  
  set_table_properties(z, width = 1, layout = "fixed")
  
  return(z)
}

# 3-way ANOVA
ft_aov3 <- function(x, end_title){
  
  z <- flextable(x) %>% 
  # force specific number of decimal places
    set_formatter(
      `num Df` = function(x) sprintf("%.1f", x),
      `den Df` = function(x) sprintf("%.1f", x),
      MSE = function(x) sprintf("%.2f", x),
      `F` = function(x) sprintf("%.2f", x),
      ges = function(x) sprintf("%.3f", x),
      p = function(x) sprintf("%.4f", x)) %>% 
  # more cosmetics
  # set_header_labels() %>% 
    add_footer_lines("* p<0.05, ** p<0.01, *** p<0.001, **** p<0.0001") %>% 
    color(part = "footer", color = "#666666") %>% 
  # custom title
    set_caption(as_paragraph(
      as_chunk(paste0("Three-way ANOVA of ", end_title),
               props = fp_text_default(font.family = "Arial")))) %>% 
    # custom width for first column
    width(j = 1, width = 1.2)
  
  set_table_properties(z, width = 1, layout = "fixed")
  
  return(z)
}

# 3-way ANOVA post hoc
ft_aov3_post <- function(x, end_title){
  
  z <- flextable(x) %>% 
  # force specific number of decimal places
    set_formatter(
      estimate = function(x) sprintf("%.3f", x),
      SE = function(x) sprintf("%.3f", x),
      t.ratio = function(x) sprintf("%.4f", x),
      p.value = function(x) sprintf("%.4f", x),
      p.adj = function(x) sprintf("%.4f", x)) %>% 
  # add p symbols as footer
    add_footer_lines("* p<0.05, ** p<0.01, *** p<0.001, **** p<0.0001") %>% 
    color(part = "footer", color = "#666666") %>% 
  # custom title
    set_caption(as_paragraph(
      as_chunk(paste0("Post hoc comparisons from three-way ANOVA of ", end_title),
               props = fp_text_default(font.family = "Arial")))) %>%
  # custom width for first column
    width(j = 1, width = 1.2)
  
    # condition cosmetics
  if ("day" %in% names(x)) {
    z <- set_header_labels(z, contrast = "Contrast", day = "Day",
                           estimate = "Estimate", SE = "SE", df = "df",
                           t.ratio = "t-ratio", p.value = "p-value", 
                           "p<.05" = "p<.05") 
    
    } else if ("session" %in% names(x)) {
      z <- set_header_labels(z, contrast = "Contrast", session = "Session",
                             estimate = "Estimate", SE = "SE", df = "df",
                             t.ratio = "t-ratio", p.value = "p-value", 
                             "p<.05" = "p<.05") 
      
      } else {z <- set_header_labels(z, names(x))} 
      
  set_table_properties(z, width = 1, layout = "fixed")
  
  return(z)
}


# add_header_row(ft,
#                colwidths = c(4, 2),
#                values = c(" ", " ")) %>%

# to format color by p-value significance
# set_formatter(values = list("p.value" = p_val_format) ) %>% 
#   color(i = ~ p.value > .05, color = "#DFDFDF") %>% 
```

```{r}
df_weight_m_aov3 <- read_csv("data/results/weight_m_aov3.csv")
df_weight_f_aov3 <- read_csv("data/results/weight_f_aov3.csv")

df_percent_weight_m_aov3 <- read_csv("data/results/percent_weight_m_aov3.csv")
df_percent_weight_f_aov3 <- read_csv("data/results/percent_weight_f_aov3.csv")

df_intox_aov3 <- read_csv("data/results/intox_aov3.csv")
df_intox_aov3_post <- read_csv("data/results/intox_aov3_post.csv")

df_daily_dose_aov3 <- read_csv("data/results/daily_dose_aov3.csv")
df_daily_dose_aov3_post <- read_csv("data/results/daily_dose_aov3_post.csv")

df_bec_aov2 <- read_csv("data/results/bec_aov2.csv")


ft1 <- ft_aov3(x = df_weight_m_aov3, end_title = "Male Body Weight (g)")
ft2 <- ft_aov3(x = df_weight_f_aov3, end_title = "Female Body Weight (g)")

ft3 <- ft_aov3(x = df_percent_weight_m_aov3, end_title = "Male Percent Body Weight")
ft4 <- ft_aov3(x = df_percent_weight_f_aov3, end_title = "Female Percent Body Weight")

ft5 <- ft_aov3(x = df_intox_aov3, end_title = "Intoxication Score")
ft6 <- ft_aov3_post(x = df_intox_aov3_post, end_title = "Intoxication Score")

ft7 <- ft_aov3(x = df_daily_dose_aov3, end_title = "Daily Ethanol Dose (g/kg/day)")
ft8 <- ft_aov3_post(x = df_daily_dose_aov3_post, end_title = "Daily Ethanol Dose (g/kg/day)")

ft9 <- ft_aov2(x = df_bec_aov2, end_title = "Blood Ethanol Concentration (mg/dl)")


save_as_docx(
  "df_weight_m_aov3" = ft1, "df_weight_f_aov3" = ft2, 
  "df_percent_weight_m_aov3" = ft3, "df_percent_weight_f_aov3" = ft4, 
  "df_intox_aov3" = ft5, "df_intox_aov3_post" = ft6, 
  "df_daily_dose_aov3" = ft7, "df_daily_dose_aov3_post" = ft8, 
  "df_bec_aov2" = ft9,
  path = "data/results/aov.docx")

```


############# BONUS ANALYSIS ###################################################

## Mixed effect model of BEC and dose
**2025-10-01**

**Need to work on model fit**

```{r mixed effect model of BEC and dose}
binge_long <- binge %>% select(subject:bec_mean, dose_01:dose_07) %>% pivot_longer(cols = dose_01:dose_07, names_to = "time", values_to = "dose")

# convert to factors, relevel drug factor
binge_long <- binge_long %>% 
  mutate(dose = factor(dose), drug = factor(drug), sex = factor(sex)) %>% 
  mutate(drug = fct_relevel(drug, c("VEH", "PLX")))

# Fit the mixed-effects model
model <- lme4::lmer(bec_mean ~ dose * drug * sex + (1|subject), data = binge_long)
# not a good fit
# fit warnings:
# fixed-effect model matrix is rank deficient so dropping 1 column / coefficient
# optimizer (nloptwrap) convergence code: 0 (OK)
# Model failed to converge with max|grad| = 1.29232 (tol = 0.002, component 1)
# Model is nearly unidentifiable: very large eigenvalue

# Summary of the model
summary(model)
```

################################################################################



# Archive

## Correlation plots of binge metrics

Code to convert drug_sex legend labels from "PLX_F" format -> "PLX female" format
to improve correlation scatterplot legend readability within plot creation

Alternative method = change format of drug_sex column values within data table

```{r}
# Create custom aesthetics
custom_colors <- c("PLX female"="#FF9933", "PLX male"="#BF5700", "VEH female"="#0000FF", "VEH male"="#000080")
custom_fills <- c("PLX female"="#FF9933", "PLX male"="#FFFFFF", "VEH female"="#0000FF", "VEH male"="#FFFFFF")
custom_shapes <- c("PLX female"=22, "PLX male"=22, "VEH female"=21, "VEH male"=21)
custom_linetypes <- c("PLX female"="solid", "PLX male"="dashed", "VEH female"="solid", "VEH male"="dashed")

# PLX female: shape=22, fill="#BF5700", color="#BF5700", orange filled square w/ orange border and solid line
# PLX male: shape=22, fill="#FFFFFF", color="#BF5700", white filled square w/ orange border and dashed line
# VEH female: shape=21, fill="#0000FF", color="#0000FF", blue filled circle with blue border and solid line
# VEH male: shape=21, fill="#FFFFFF", color="#0000FF", white filled circle w/ blue border and dashed line

# dose 7 vs BEC
plot1 <- ggplot(data = binge, 
                aes(x = dose_07, 
                    y = bec_mean)) + 
  geom_point(aes(shape = drug_sex, 
                 fill = drug_sex, 
                 color = drug_sex), 
             size = 3) + # scatterplot
  scale_shape_manual(values = custom_shapes,
                     labels = c("PLX female", "PLX male", "VEH female", "VEH male")) +
  scale_fill_manual(values = custom_fills,
                    labels = c("PLX female", "PLX male", "VEH female", "VEH male")) +
  scale_color_manual(values = custom_colors,
                     labels = c("PLX female", "PLX male", "VEH female", "VEH male")) +
  scale_linetype_manual(values = custom_linetypes,
                        labels = c("PLX female", "PLX male", "VEH female", "VEH male")) +
  geom_smooth(method = lm, formula = y ~ x, se = FALSE, 
              aes(color = drug_sex, linetype = drug_sex)) + # add line of fit
  scale_x_continuous(limits = c(0, 5), expand = c(0, 0.01)) + # set x axis bounds
  scale_y_continuous(limits = c(0, 500), expand = c(0, 1)) + # set y axis bounds
  theme_prism() + # use preset theme for prism-like graph
  ylab("BEC (mg/dl) \n90 min post-ethanol dose") +
  xlab("Ethanol dose (g/kg)") +
  labs(title = "Correlation of 7th ethanol dose and BEC") + 
  stat_cor(method = "pearson", aes(color = drug_sex, group = drug_sex), label.x.npc = 0.05, label.y.npc = 0.25, 
           show.legend = FALSE) # add correlation coefficients to graph
plot1
```
