---
title: "MGD56_Analyses"
author: "Erika Carlson"
format: html
editor_options: 
  chunk_output_type: console
---

------

This document contains research notes and code generated by:

> **Erika R. Carlson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> erika.carlson@utexas.edu

With assistance from:

> **K. Ryan Thompson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> kt25399@utexas.edu

------

# Overview

We examined the role of microglia in alcohol-induced neurodegeneration in a rat model of an AUD by depleting microglia with PLX5622 (PLX), a highly selective antagonist of colony-stimulating factor-1 receptor (CSF1R), a receptor kinase critical for survival and proliferation of microglia/macrophages.

Iba1 (ionized calcium binding adapter 1) is a marker of microglia/macrophages.
FJB (FluoroJade-B) is a marker of dying neurons.

Adult female and male sprague-dawley rats were injected with PLX [50 mg/kg, 10 mL/kg, subcutaneous (s.c.), every 12 h] or vehicle for 11 days. During the last 4 days of PLX, rats received ethanol (0-5 g/kg, 25% EtOH w/v, intragastric (i.g.), every 8 h) in diluted Vanilla Ensure Plus (50% w/v) in a Majchrowicz paradigm. Tissue was collected during intoxication, 3-6 h post-EtOH (T0).
Blood ethanol concentration (BEC; mg/dL) was determined from tail blood collected 90 min post-dose 7.
Tissue was prepared by phosphate-buffered saline perfusion (PBS), brain extraction and bisection, at least 24 h post-fixation in 4% paraformaldehyde (PFA) at 4°C, and storage in PBS at 4°C until sliced 1:12 series at 40 um thickness on vibrating microtome. Tissue slices were stored in cryoprotectant at -20°C until immunohistochemical assay.



# Preliminaries

- Load packages
- Indicate path to save results

```{r}
#| include: false
#| message: false
library(tidyverse)
library(rstatix) # for statistical analysis
library(emmeans) # for statistical analysis
library(lme4) # for statistical analysis
library(afex) # dependency for ANOVA analysis
library(ggplot2) # for data visualization
library(ggpubr) # for data visualization

# Indicate path to save results
data_path <- "data/"
results_path <- "data/results/"
```


## Function to add significance to {afex} ANOVA results

`sig_sym()` is a function that adds a significance symbol column to an ANOVA 
table and saves the results as a tibble.

It has the following arguments: 

- **aov_res** is a model object of class list containing results of `aov_ez()`
- **path** indicates the destination path to save files. 
  - It can be a named vector of class character or a quoted path e.g., "data/results/"
  
The function does the following:

- Creates **name_aov_res**, a string, from input **aov_res**, a model object list.
  - This will be used to name the output file
- Format the ANOVA results into a table
  - Compute ANOVA tables for the input model object **aov_res**
  - Convert row names containing the model factors to a column with name Factor
  - Convert data frame to a tibble
  - Rename **Pr(>F)** column containing p-values to **p**
  - Add new column **p<.05** with significance symbols for values of **p**
  - Save results as ".csv". 
    - Path indicated by **path** and **name_aov_res**
    
Dependencies: {tidyverse}

```{r}
aov_sig_sym <- function(aov_res, path) {
  
  # Create object of aov_res in quotes to name file
  name_aov_res <- deparse(substitute(aov_res))
  
  # Format ANOVA table, add significance symbol column, save
  res <- anova(aov_res) %>%
  rownames_to_column(var = "Factor") %>%
  as_tibble() %>% 
  rename(p = "Pr(>F)") %>% 
  mutate("p<.05" = case_when(p < 0.0001 ~ "****",
                             p < 0.001 ~ "***",
                             p < 0.01 ~ "**",
                             p < 0.05 ~ "*",
                             p >= 0.05 ~ "ns")) %>% 
  write_csv(., str_c(path, name_aov_res, ".csv"))
  
  return(res)
}
```


## Function to add significance to {afex} emmeans results

`emm_sig_sym()` is a function that adds a significance symbol column to an 
emmeans contrasts tibble and saves the results as a ".csv".

It has the following arguments: 

- **emm_res** is a tibble containing contrasts from `emmeans()` model object
- **path** indicates the destination path to save files. 
  - It can be a named vector of class character or a quoted path e.g., "data/results/"
  
The function does the following:

- Creates **name_emm_res**, a string, from input **emm_res**, 
a tibble containing contrasts from `emmeans()` model object.
  - This will be used to name the output file
- Format the emmeans results into a table
  - Convert data frame to a tibble (redundant)
  - Add new column **p<.05** with significance symbols for values of **p**
  - Save results as ".csv". 
    - Path indicated by **path** and **name_emm_res**
    
Dependencies: {tidyverse}

```{r}
emm_sig_sym <- function(emm_res, path) {
  
  # Create object of emm_res in quotes to name file
  name_emm_res <- deparse(substitute(emm_res))
  
  # Format emmeans table, add significance symbol column, save
  res <- emm_res %>%
  as_tibble() %>%
  mutate("p<.05" = case_when(p.adj < 0.0001 ~ "****",
                             p.adj < 0.001 ~ "***",
                             p.adj < 0.01 ~ "**",
                             p.adj < 0.05 ~ "*",
                             p.adj >= 0.05 ~ "ns")) %>% 
  write_csv(., str_c(path, name_emm_res, ".csv"))
  
  return(res)
}
```

## Function to extract ANOVA results sentence from {afex} anova model object

`add_anova_summary()` is a function to extract anova sentence and append to *anova_summary_df*

It has the following arguments: 

- **aov_res** is a model object of class list containing results of `aov_ez()`
- **data_name** is the quoted name of the data set to analyze (e.g., "measure_region")

Dependencies: {dplyr}, {glue}, {afex}

```{r}
# Dependencies
# library(dplyr)
# library(glue)
# library(afex)

# Function to extract sentence and append to **anova_summary_df**
add_anova_summary <- function(aov_res, data_name) {
  
  # Extract and format ANOVA sentence
  anova_tbl <- anova(aov_res) %>%
    rownames_to_column(var = "Factor") %>%
    as_tibble()

  anova_sentence <- anova_tbl %>%
    mutate(
      F_val = sprintf("%.2f", `F`),
      ges_val = sprintf("%.3f", `ges`),
      p_val = sprintf("%.4f", `Pr(>F)`),
      num.Df = sprintf("%.0f", `num Df`),
      den.Df = sprintf("%.0f", `den Df`)
      ) %>%
    transmute(
      text = glue("{Factor} (F{num.Df},{den.Df} = {F_val}, \u03B72G = {ges_val}, p = {p_val})")
      ) %>%
    pull(text) %>%
    paste(collapse = ", ") %>%
    paste0("ANOVA yielded a main effect of ", .)

  # Capture name of anova object
  aov_name <- deparse(substitute(aov_res))
  
   # Create new row
  new_row <- tibble(
    data_frame = data_name,
    anova_object = aov_name,
    anova_sentence = anova_sentence
  )

  # Create or append to summary df
  if (!exists("anova_summary_df", envir = .GlobalEnv)) {
    assign("anova_summary_df", new_row, envir = .GlobalEnv)
  } else {
    anova_summary_df <<- bind_rows(anova_summary_df, new_row)
  }

  # Return current state
  return(anova_summary_df)
}
```

```{r}
#| eval: false

# Example usage
aov3 <- aov_ez(id = 'subject', 
               dv = "hippo", 
               data = df, 
               between = c('diet', 'drug', 'sex'))


anova_summary_df <- add_anova_summary(aov_res = aov3,
                                      data_name = "fjb_hippo")
```


## Function for custom theme for {ggplot2}

```{r Graph theme}
theme_prism <- function() {
  theme_minimal(base_size = 16) +  # Use a minimal base theme
    theme(
      panel.grid.major = element_blank(),  # No major grid lines
      panel.grid.minor = element_blank(),  # No minor grid lines
      panel.background = element_rect(fill = "white", color = NA),  # White background
      plot.background = element_rect(fill = "white", color = NA),  # White plot background
      axis.text = element_text(color = "black"),  # Black axis text
      axis.title = element_text(color = "black"),  # Black axis titles
      axis.ticks = element_line(color = "black"),  # Black ticks
      axis.line.x = element_line(color = "black"),  # Bottom axis line
      axis.line.y = element_line(color = "black"),  # Left axis line
      legend.position = "right",  # Legend position in x
      legend.justification = "top", # Legend position in y
      legend.background = element_blank(),  # Transparent legend background
      legend.key = element_blank(),  # Remove outline around legend key
      legend.title = element_blank(), # Remove legend key title
      plot.title = element_text(hjust = 0.5, size = 16),  # Centered title
      plot.subtitle = element_text(hjust = 0.5, size = 12)  # Centered subtitle
    )
}
```


# Binge statistics reformatting 

Combining MGD5 and MGD6 data

```{r}
# Load dose data, retaining excluded subjects
dose <- read_csv("data/dose.csv") # %>% filter(exclude == "N")

# Check data
slice_sample(dose, n = 6)

# Calculate average dose per day per subject in daily_dose table
daily_dose <- dose %>% 
  mutate(day_01 = rowSums(select(., dose_01:dose_03)),
         day_02 = rowSums(select(., dose_04:dose_06)),
         day_03 = rowSums(select(., dose_07:dose_09)),
         day_04 = rowSums(select(., dose_10:dose_12))) %>% 
  select(!dose_01:dose_12)

write_csv(daily_dose, "data/daily_dose.csv")
```

## Custom function

`rearrange_data` function to reorder data by factor columns

```{r}
# Function to rearrange data and resave file
# Rearrange as sex: F, M - drug: VEH, PLX - diet: CON, EtOH

rearrange_data <- function(path, data, cols,
                           desc_cols = NULL) {
  
  # Load data
  df <- read_csv(str_c(path, data, ".csv"))
  
  # Build list of order expressions, applying desc() where specified
  order_exprs <- lapply(cols, function(col) {
    if (col %in% desc_cols) desc(df[[col]]) else df[[col]]
  })
  
  # Rearrange by columns in order_exprs, unlisted with splice operator
  df <- df %>% arrange(!!!order_exprs)
  
  # Save ".csv"
  write_csv(df, str_c(path, data, ".csv"))
  
  # Check
  head(df)
}
```

```{r}
# Non-function way
data <- read_csv("data/percent_baseline_weight.csv")
# rearrange as sex: F, M - drug: VEH, PLX - diet: CON, EtOH
data <- data %>% arrange(sex, desc(drug), diet)
# save .csv
write_csv(data, "data/percent_baseline_weight.csv")
```

```{r}
# Weight
rearrange_data(data_path, data = "weight", 
               cols = c("sex", "drug", "diet"), 
               desc_cols = c("drug"))

# Weight, percent baseline
rearrange_data(data_path, data = "percent_baseline_weight", 
               cols = c("sex", "drug", "diet"), 
               desc_cols = c("drug"))

# Intoxication
rearrange_data(data_path, data = "intoxication", 
               cols = c("drug", "sex"), 
               desc_cols = c("drug"))

# Dose
rearrange_data(data_path, data = "dose", 
               cols = c("drug", "sex"), 
               desc_cols = c("drug"))

# Daily dose
rearrange_data(data_path, data = "daily_dose", 
               cols = c("drug", "sex"), 
               desc_cols = c("drug"))

# BEC
rearrange_data(data_path, data = "bec", 
               cols = c("drug", "sex"), 
               desc_cols = c("drug"))
```


# Binge statistics

**2026-04-09**
Run on MGD5 and MGD6

Excluding rats that died during binge: 17, 36, 31, 28, 52, 67, 78

May later exclude #35 (hind limb/paw edema) <- would need to update exclude column in all data files

```{r}
results_path <- "data/results/"
```

```{r Load and process body weights}
# Load weight data
weight <- read_csv("data/weight.csv") %>% filter(exclude == "N")

# Check data
slice_sample(weight, n = 6)

# Get summary statistics per day per sex
weight_common <- weight %>% 
  group_by(sex) %>% 
  get_summary_stats(day_01:day_11, type = "common") %>% 
  as_tibble() %>% 
  write_csv(., str_c(results_path, "weight_summary.csv"))

# Obtain mean ± se onset weight overall
weight_onset <- weight %>% 
  group_by(sex) %>% 
  get_summary_stats(day_01, type = "mean_se") %>% 
  as_tibble() %>% 
  write_csv(., str_c(results_path, "weight_onset.csv"))


# Compute two-way ANOVA on average onset weight in males using {afex} package
weight_onset_m_aov2 <- aov_ez(id = 'subject', 
                              dv = 'day_01', 
                              data = weight %>% filter(sex == 'M'), 
                              between = c('diet', 'drug')
                              )

# Add significance symbol and save result as tibble using custom `sig_sym()` function
aov_sig_sym(weight_onset_m_aov2, results_path)
# no significance

# Create ANOVA sentence using custom `add_anova_summary()` function
add_anova_summary(weight_onset_m_aov2, "weight_onset_m")



# Repeat in females
weight_onset_f_aov2 <- aov_ez(id = 'subject', 
                              dv = 'day_01', 
                              data = weight %>% filter(sex == 'F'), 
                              between = c('diet', 'drug')
                              )

aov_sig_sym(weight_onset_f_aov2, results_path)
# no significance

add_anova_summary(weight_onset_f_aov2, "weight_onset_f")



# Create weight_long table for analysis across days
weight_long <- weight %>% pivot_longer(cols = day_01:day_11, names_to = "day", values_to = "weight")

# Compute three-way ANOVA on average weight over time in males using {afex} package
weight_m_aov3 <- aov_ez(id = 'subject', 
                        dv = 'weight', 
                        data = weight_long %>% filter(sex == 'M'), 
                        between = c('diet', 'drug'),
                        within = 'day'
                              )

aov_sig_sym(weight_m_aov3, results_path)
# main effect of drug, day
# interaction of diet x day, drug x day, diet x drug x day

add_anova_summary(weight_m_aov3, "weight_m")

# Post-hoc comparisons, restrict to common factors (no interaction)
weight_m_aov3_post <- emmeans(weight_m_aov3, 
                           pairwise ~ diet * drug | day,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("CON PLX - EtOH PLX",
                         "CON PLX - CON VEH", 
                         "EtOH PLX - EtOH VEH",
                         "CON VEH - EtOH VEH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni")) # corrected for 44 comparisons

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(weight_m_aov3_post, results_path)


# Repeat in females
weight_f_aov3 <- aov_ez(id = 'subject', 
                        dv = 'weight', 
                        data = weight_long %>% filter(sex == 'F'), 
                        between = c('diet', 'drug'),
                        within = 'day'
                              )

aov_sig_sym(weight_f_aov3, results_path)
# main effect of day
# interaction of drug x day, diet x day

add_anova_summary(weight_f_aov3, "weight_f")


# Post-hoc comparisons, restrict to common factors (no interaction)
weight_f_aov3_post <- emmeans(weight_f_aov3, 
                           pairwise ~ diet * drug | day,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("CON PLX - EtOH PLX",
                         "CON PLX - CON VEH", 
                         "EtOH PLX - EtOH VEH",
                         "CON VEH - EtOH VEH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni")) # corrected for 44 comparisons

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(weight_f_aov3_post, results_path)


###############################################################################


# Load % baseline weight data
percent_weight <- read_csv("data/percent_baseline_weight.csv") %>% filter(exclude == "N")

# Check data
slice_sample(percent_weight, n = 6)

# Create percent_weight_long table for analysis across days
percent_weight_long <- percent_weight %>% pivot_longer(cols = day_01:day_11, names_to = "day", values_to = "weight")


# Compute three-way ANOVA on average weight over time in males using {afex} package
percent_weight_m_aov3 <- aov_ez(id = 'subject', 
                        dv = 'weight', 
                        data = percent_weight_long %>% filter(sex == 'M'), 
                        between = c('diet', 'drug'),
                        within = 'day'
                              )

# Use custom function `sig_sym()` to add significance symbol and save result as tibble
aov_sig_sym(percent_weight_m_aov3, results_path)
# main effect of drug, day (diet p = 0.051)
# interaction of diet x day, drug x day, diet x drug x day

add_anova_summary(percent_weight_m_aov3, "percent_weight_m")



# Repeat in females
percent_weight_f_aov3 <- aov_ez(id = 'subject', 
                        dv = 'weight', 
                        data = percent_weight_long %>% filter(sex == 'F'), 
                        between = c('diet', 'drug'),
                        within = 'day'
                              )

aov_sig_sym(percent_weight_f_aov3, results_path)
# main effect of drug, day (diet p = 0.078)
# interaction of drug x day, diet x day

add_anova_summary(percent_weight_f_aov3, "percent_weight_f")
```


```{r Load and process intoxication scores}
# Load behavioral intoxication score data
intox <- read_csv("data/intoxication.csv") %>% filter(exclude == "N")

# Check data
slice_sample(intox, n = 6)

# Calculate average behavioral intoxication score per subject
intox <- intox %>%
  mutate(mean = rowMeans(select(., score_01:score_12)))

# Get summary statistics
intox_summary <- intox %>%
  group_by(drug, sex) %>%
  get_summary_stats(mean, type = "mean_se") %>% 
  as_tibble() %>% 
  write_csv(., str_c(results_path, "/intoxication_summary.csv"))
intox_summary


# Compute two-way ANOVA on average behavioral intoxication using {afex} package
intox_aov2 <- aov_ez(id = 'subject', 
                     dv = 'mean', 
                     data = intox, 
                     between = c('sex', 'drug')
                     )

# Use custom function `sig_sym()` to add significance label and save result as tibble
aov_sig_sym(intox_aov2, results_path)
# Main effect of sex, drug
# No interaction

add_anova_summary(intox_aov2, "intox")


# Reformat data for three-way ANOVA across dose sessions (time)
intox_long <- intox %>% select(!mean) %>% 
  pivot_longer(cols = score_01:score_12, 
               names_to = "session",
               values_to = "score")

# Compute three-way ANOVA on behavioral intoxication across time using {afex} package
intox_aov3 <- aov_ez(id = 'subject', 
                     dv = 'score', 
                     data = intox_long, 
                     between = c('sex', 'drug'),
                     within = 'session'
                     )

intox_aov3

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(intox_aov3, results_path)
# Main effect of sex, drug, session
# sex x session, drug x session interaction

add_anova_summary(intox_aov3, "intox")


# Post-hoc comparisons, could restrict to common factors (no interaction) with $contrasts[c()]
# within session: [1] F PLX - M PLX [2] F PLX - F VEH [3] F PLX - M VEH [4] M PLX - F VEH [5] M PLX - M VEH [6] F PLX - M PLX
# 72 total comparisons reduced to 44 (those that share a factor level)
# Note: score_01 skipped (all rats score 0)
# Note: session are repeated measures, correct across all sessions (outside of emmeans())
intox_aov3_post <- emmeans(intox_aov3, 
                           pairwise ~ sex * drug | session,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(session != "score_01") %>% 
  filter(contrast %in% c("F PLX - M PLX",
                         "F VEH - M VEH", 
                         "F PLX - F VEH",
                         "M PLX - M VEH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni")) # corrected for 44 comparisons

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(intox_aov3_post, results_path)

# Sig "main effect" diffs: score_04 F PLX - M PLX, score_10 F PLX - M PLX, score_10 F PLX - F VEH, score_11 F PLX - F VEH
# also some interaction diffs: M PLX - F VEH, F PLX - M VEH
```

```{r Load and process daily dose}
# Load dose data
dose <- read_csv("data/dose.csv") %>% filter(exclude == "N")

# Load daily dose data
daily_dose <- read_csv("data/daily_dose.csv") %>% filter(exclude == "N")

# Check data
slice_sample(daily_dose, n = 6)

# Calculate average daily dose per subject
daily_dose <- daily_dose %>%
  mutate(mean = rowMeans(select(., day_01:day_04)))


# Get summary statistics
daily_dose_summary <- daily_dose %>%
  group_by(drug, sex) %>%
  get_summary_stats(mean, type = "mean_se") %>% 
  as_tibble() %>% 
  write_csv(., str_c(results_path, "/daily_dose_summary.csv"))
daily_dose_summary


# Compute two-way ANOVA on average daily dose using {afex} package
daily_dose_aov2 <- aov_ez(id = 'subject', 
                     dv = 'mean', 
                     data = daily_dose, 
                     between = c('sex', 'drug')
                     )
daily_dose_aov2


# Use custom function `sig_sym()` to add significance label and save result as tibble
aov_sig_sym(daily_dose_aov2, results_path)
# Main effect of sex, drug
# No interaction

add_anova_summary(daily_dose_aov2, "daily_dose")


# Reformat data for three-way ANOVA across days (time)
daily_dose_long <- daily_dose %>% select(!mean) %>% 
  pivot_longer(cols = day_01:day_04, 
               names_to = "day",
               values_to = "daily_dose")

# Compute three-way ANOVA on daily dose across days using {afex} package
daily_dose_aov3 <- aov_ez(id = 'subject', 
                     dv = 'daily_dose', 
                     data = daily_dose_long, 
                     between = c('sex', 'drug'),
                     within = 'day'
                     )

# Use custom function `aov_sig_sym()` to add significance label and save result as tibble
aov_sig_sym(daily_dose_aov3, results_path)
# Main effect of sex, drug, day
# No interactions (sex x day p = 0.06)

add_anova_summary(daily_dose_aov3, "daily_dose")


# Post-hoc comparisons, restrict to common factors (no interaction)
# within day: [1] F PLX - M PLX [2] F PLX - F VEH [3] F PLX - M VEH [4] M PLX - F VEH [5] M PLX - M VEH [6] F PLX - M PLX
# 24 possible comparisons total
daily_dose_aov3_post <- emmeans(daily_dose_aov3, 
                           pairwise ~ sex * drug | day,
                           adjust = "none")$contrasts %>% 
  as_tibble() %>% 
  filter(contrast %in% c("F PLX - M PLX",
                         "F VEH - M VEH", 
                         "F PLX - F VEH",
                         "M PLX - M VEH")) %>% 
  mutate(p.adj = p.adjust(p.value, method = "bonferroni")) # corrected for 16 comparisons

# Use custom function `emm_sig_sym()` to add significance label and save result as tibble
emm_sig_sym(daily_dose_aov3_post, results_path)

# main effect diffs: day_02 F PLX - M PLX
```


```{r Load and process blood ethanol concentration}
# Load blood ethanol concentration (bec) data
bec <- read_csv("data/bec.csv") %>% filter(exclude == "N")

# Check data
slice_sample(bec, n = 6)

# Get summary statistics
bec_summary <- bec %>%
  group_by(drug, sex) %>%
  get_summary_stats(mean, type = "mean_se") %>% 
  as_tibble() %>% 
  write_csv(., str_c(results_path, "/bec_summary.csv"))
bec_summary


# Compute two-way ANOVA on bec using {afex} package
bec_aov2 <- aov_ez(id = 'subject', 
                     dv = 'mean', 
                     data = bec, 
                     between = c('sex', 'drug')
                     )
bec_aov2
# No sig effects

# Use custom function `sig_sym()` to add significance label and save result as tibble
aov_sig_sym(bec_aov2, results_path)

add_anova_summary(bec_aov2, "bec")
```

```{r Correlate bec and dose 7, bec and mean intox, bec and mean daily dose}
# bind bec and dose data frames (bec$mean renamed to bec_mean)
binge <- full_join(bec %>% select(subject, diet, drug, sex, mean) %>% rename(bec_mean = mean), 
                      dose %>% filter(exclude == "N") %>% 
                        select(subject, diet, drug, sex, dose_01:dose_12))

# bind daily_dose, including daily_dose$mean (renamed to daily_dose_mean) to binge data frame
binge <- full_join(binge, daily_dose %>% select(subject, diet, drug, sex, mean) %>% rename(daily_dose_mean = mean))

# bind intox, including intox$mean (renamed to intox_mean) to binge data frame
binge <- full_join(binge, intox %>% select(subject, diet, drug, sex, mean) %>% rename(intox_mean = mean))

# rename F to female and M to male (better for graph legend)
binge <- binge %>% mutate(sex = case_when(sex == 'M' ~ 'male',
                                 sex == 'F' ~ 'female',
                                 TRUE ~ sex))

# move mean columns to after bec_mean and create drug_sex column 
binge <- binge %>% relocate(daily_dose_mean:intox_mean, .after = bec_mean) %>% 
  mutate(drug_sex = paste(drug, sex, sep = " "), .before = bec_mean)

# check order of drug_sex factor levels
binge$drug_sex <- as.factor(binge$drug_sex) 


# Calculate correlation between ethanol dose 7 and BEC
cor_bec_dose <- binge %>% group_by(drug, sex) %>% cor_test(dose_07, bec_mean)


# Calculate correlation between mean intoxication and BEC
cor_bec_intox <- binge %>% group_by(drug, sex) %>% cor_test(intox_mean, bec_mean)

# Calculate correlation between mean daily dose and BEC
cor_bec_dd <- binge %>% group_by(drug, sex) %>% cor_test(daily_dose_mean, bec_mean)


# Create custom aesthetics
custom_colors <- c("PLX female"="#FF9933", "PLX male"="#BF5700", "VEH female"="#0000FF", "VEH male"="#000080")
custom_fills <- c("PLX female"="#FF9933", "PLX male"="#FFFFFF", "VEH female"="#0000FF", "VEH male"="#FFFFFF")
custom_shapes <- c("PLX female"=22, "PLX male"=22, "VEH female"=21, "VEH male"=21)
custom_linetypes <- c("PLX female"="solid", "PLX male"="dashed", "VEH female"="solid", "VEH male"="dashed")

# PLX female: shape=22, fill="#BF5700", color="#BF5700", orange filled square w/ orange border and solid line
# PLX male: shape=22, fill="#FFFFFF", color="#BF5700", white filled square w/ orange border and dashed line
# VEH female: shape=21, fill="#0000FF", color="#0000FF", blue filled circle with blue border and solid line
# VEH male: shape=21, fill="#FFFFFF", color="#0000FF", white filled circle w/ blue border and dashed line

# dose 7 vs BEC
plot1 <- ggplot(data = binge, 
                aes(x = dose_07, 
                    y = bec_mean)) + 
  geom_point(aes(shape = drug_sex, 
                 fill = drug_sex, 
                 color = drug_sex), 
             size = 3) + # scatterplot
  scale_shape_manual(values = custom_shapes) +
  scale_fill_manual(values = custom_fills) +
  scale_color_manual(values = custom_colors) +
  scale_linetype_manual(values = custom_linetypes) +
  geom_smooth(method = lm, formula = y ~ x, se = FALSE, 
              aes(color = drug_sex, linetype = drug_sex)) + # add line of fit
  scale_x_continuous(limits = c(0, 5), expand = c(0, 0.01)) + # set x axis bounds
  scale_y_continuous(limits = c(0, 500), expand = c(0, 1)) + # set y axis bounds
  theme_prism() + # use preset theme for prism-like graph
  ylab("BEC (mg/dl) \n90 min post-ethanol dose") +
  xlab("Ethanol dose (g/kg)") +
  labs(title = "Correlation of 7th ethanol dose and BEC") + 
  stat_cor(method = "pearson", aes(color = drug_sex, group = drug_sex), label.x.npc = 0.05, label.y.npc = 0.25, 
           show.legend = FALSE) # add correlation coefficients to graph
plot1

# save plot
ggsave("corr_BECvDose7.tiff", path = "data/graphs/", width = 6.5, height = 4.5, units = "in")


# mean intox vs BEC
plot2 <- ggplot(data = binge, 
                aes(x = intox_mean, 
                    y = bec_mean)) + 
  geom_point(aes(shape = drug_sex, 
                 fill = drug_sex, 
                 color = drug_sex), 
             size = 3) + # scatterplot
  scale_shape_manual(values = custom_shapes) +
  scale_fill_manual(values = custom_fills) +
  scale_color_manual(values = custom_colors) +
  scale_linetype_manual(values = custom_linetypes) +
  geom_smooth(method = lm, formula = y ~ x, se = FALSE, 
              aes(color = drug_sex, linetype = drug_sex)) + # add line of fit
  scale_x_continuous(limits = c(0, 5), expand = c(0, 0.01)) + # set x axis bounds
  scale_y_continuous(limits = c(0, 500), expand = c(0, 1)) + # set y axis bounds
  theme_prism() + # use preset theme for prism-like graph
  ylab("BEC (mg/dl)") +
  xlab("Mean intoxication score") +
  labs(title = "Correlation of mean intoxication and BEC") + 
  stat_cor(method = "pearson", aes(color = drug_sex, group = drug_sex), label.x.npc = 0.05, label.y.npc = 0.25, 
           show.legend = FALSE) # add correlation coefficients to graph
plot2

ggsave("corr_BECvIntox.tiff", path = "data/graphs/", width = 6.5, height = 4.5, units = "in")


# mean daily dose vs BEC
plot3 <- ggplot(data = binge, 
                aes(x = daily_dose_mean, 
                    y = bec_mean)) + 
  geom_point(aes(shape = drug_sex, 
                 fill = drug_sex, 
                 color = drug_sex), 
             size = 3) + # scatterplot
  scale_shape_manual(values = custom_shapes) +
  scale_fill_manual(values = custom_fills) +
  scale_color_manual(values = custom_colors) +
  scale_linetype_manual(values = custom_linetypes) +
  geom_smooth(method = lm, formula = y ~ x, se = FALSE, 
              aes(color = drug_sex, linetype = drug_sex)) + # add line of fit
  scale_x_continuous(limits = c(0, 15), expand = c(0, 0.01)) + # set x axis bounds
  scale_y_continuous(limits = c(0, 500), expand = c(0, 1)) + # set y axis bounds
  theme_prism() + # use preset theme for prism-like graph
  ylab("BEC (mg/dl)") +
  xlab("Mean daily ethanol dose (g/kg/day)") +
  labs(title = "Correlation of mean daily ethanol dose and BEC") + 
  stat_cor(method = "pearson", aes(color = drug_sex, group = drug_sex), label.x.npc = 0.05, label.y.npc = 0.25, 
           show.legend = FALSE) # add correlation coefficients to graph
plot3

ggsave("corr_BECvDailyDose.tiff", path = "data/graphs/", width = 6.5, height = 4.5, units = "in")
```


############# RESULTS TABLES in "MGD56_ResultsTables.qmd" ######################

```{r}
flextable(anova_summary_df) %>% save_as_docx(., path = "data/results/anova_summary_df.docx")
```

############# BONUS ANALYSIS ###################################################

## Mixed effect model of BEC and dose
**2025-10-01**

**Need to work on model fit**

```{r mixed effect model of BEC and dose}
binge_long <- binge %>% select(subject:bec_mean, dose_01:dose_07) %>% pivot_longer(cols = dose_01:dose_07, names_to = "time", values_to = "dose")

# convert to factors, relevel drug factor
binge_long <- binge_long %>% 
  mutate(dose = factor(dose), drug = factor(drug), sex = factor(sex)) %>% 
  mutate(drug = fct_relevel(drug, c("VEH", "PLX")))

# Fit the mixed-effects model
model <- lme4::lmer(bec_mean ~ dose * drug * sex + (1|subject), data = binge_long)
# not a good fit
# fit warnings:
# fixed-effect model matrix is rank deficient so dropping 1 column / coefficient
# optimizer (nloptwrap) convergence code: 0 (OK)
# Model failed to converge with max|grad| = 1.29232 (tol = 0.002, component 1)
# Model is nearly unidentifiable: very large eigenvalue

# Summary of the model
summary(model)
```

################################################################################



# Archive

## Correlation plots of binge metrics

Code to convert drug_sex legend labels from "PLX_F" format -> "PLX female" format
to improve correlation scatterplot legend readability within plot creation

Alternative method = change format of drug_sex column values within data table

```{r}
# Create custom aesthetics
custom_colors <- c("PLX female"="#FF9933", "PLX male"="#BF5700", "VEH female"="#0000FF", "VEH male"="#000080")
custom_fills <- c("PLX female"="#FF9933", "PLX male"="#FFFFFF", "VEH female"="#0000FF", "VEH male"="#FFFFFF")
custom_shapes <- c("PLX female"=22, "PLX male"=22, "VEH female"=21, "VEH male"=21)
custom_linetypes <- c("PLX female"="solid", "PLX male"="dashed", "VEH female"="solid", "VEH male"="dashed")

# PLX female: shape=22, fill="#BF5700", color="#BF5700", orange filled square w/ orange border and solid line
# PLX male: shape=22, fill="#FFFFFF", color="#BF5700", white filled square w/ orange border and dashed line
# VEH female: shape=21, fill="#0000FF", color="#0000FF", blue filled circle with blue border and solid line
# VEH male: shape=21, fill="#FFFFFF", color="#0000FF", white filled circle w/ blue border and dashed line

# dose 7 vs BEC
plot1 <- ggplot(data = binge, 
                aes(x = dose_07, 
                    y = bec_mean)) + 
  geom_point(aes(shape = drug_sex, 
                 fill = drug_sex, 
                 color = drug_sex), 
             size = 3) + # scatterplot
  scale_shape_manual(values = custom_shapes,
                     labels = c("PLX female", "PLX male", "VEH female", "VEH male")) +
  scale_fill_manual(values = custom_fills,
                    labels = c("PLX female", "PLX male", "VEH female", "VEH male")) +
  scale_color_manual(values = custom_colors,
                     labels = c("PLX female", "PLX male", "VEH female", "VEH male")) +
  scale_linetype_manual(values = custom_linetypes,
                        labels = c("PLX female", "PLX male", "VEH female", "VEH male")) +
  geom_smooth(method = lm, formula = y ~ x, se = FALSE, 
              aes(color = drug_sex, linetype = drug_sex)) + # add line of fit
  scale_x_continuous(limits = c(0, 5), expand = c(0, 0.01)) + # set x axis bounds
  scale_y_continuous(limits = c(0, 500), expand = c(0, 1)) + # set y axis bounds
  theme_prism() + # use preset theme for prism-like graph
  ylab("BEC (mg/dl) \n90 min post-ethanol dose") +
  xlab("Ethanol dose (g/kg)") +
  labs(title = "Correlation of 7th ethanol dose and BEC") + 
  stat_cor(method = "pearson", aes(color = drug_sex, group = drug_sex), label.x.npc = 0.05, label.y.npc = 0.25, 
           show.legend = FALSE) # add correlation coefficients to graph
plot1
```
