---
title: "Slide_blinding"
author: "Erika Carlson"
date: "2025-12-30"
format: html
editor_options: 
  chunk_output_type: console
---

------

This document contains research notes and code generated by:

> **Erika R. Carlson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> erika.carlson@utexas.edu

------



# Overview

We examined the role of microglia in alcohol-induced neurodegeneration in a rat model of an AUD by depleting microglia with PLX5622 (PLX), a highly selective antagonist of colony-stimulating factor-1 receptor (CSF1R), a receptor kinase critical for survival and proliferation of microglia/macrophages.

Iba1 (ionized calcium binding adapter 1) is a marker of microglia/macrophages.
FJB (FluoroJade-B) is a marker of dying neurons.

Adult female and male sprague-dawley rats were injected with PLX [50 mg/kg, 10 mL/kg, subcutaneous (s.c.), every 12 h] or vehicle for 11 days. During the last 4 days of PLX, rats received ethanol (0-5 g/kg, 25% EtOH w/v, intragastric (i.g.), every 8 h) in diluted Vanilla Ensure Plus (50% w/v) in a Majchrowicz paradigm. Tissue was collected during intoxication, 3-6 h post-EtOH (T0).
Blood ethanol concentration (BEC; mg/dL) was determined from tail blood collected 90 min post-dose 7.
Tissue was prepared by phosphate-buffered saline perfusion (PBS), brain extraction and bisection, at least 24 h post-fixation in 4% paraformaldehyde (PFA) at 4°C, and storage in PBS at 4°C until sliced 1:12 series at 40 um thickness on vibrating microtome. Tissue slices were stored in cryoprotectant at -20°C until histological assay.

1:12 series of vibratome-cut (40 um) slices were subjected to IHC for:
- Iba1 DAB, imaged at 20X (4X representative)
- FJB, surveyed by hand, representative staining imaged at 40X

Images were blinded for analysis (then unblinded) using the following code.



# Preliminaries

- Load {tidyverse}

```{r}
#| include: false
#| message: false
library(tidyverse) # for data wrangling

# Indicate path to save results
results_path <- "data/results/"
```



# Blinding and unblinding code

## 1. Function to create a folder if it does not already exist
```{r}
create_folder <- function(folder_path) {
  # Check if the folder exists
  if (!dir.exists(folder_path)) {
    # Create the folder if it doesn't exist
    dir.create(folder_path, recursive = TRUE)  # 'recursive = TRUE' allows creation of parent directories if needed
    cat("Folder created at:", folder_path, "\n")
  } else {
    cat("Folder already exists at:", folder_path, "\n")
  }
}

# Example usage
# create_folder("path/to/your/folder")
```


## 2. Blinding subjects pre-imaging

Write `blind_subjects()` function to create blind code for slides (pre-imaging)

```{r}
# Function to create blind code for subjects
# `subjects` is a .csv with subject meta data
# `seed` is for replicability
# `subsample` used to restrict blind code creation
blind_subjects <- function(subjects, seed = 1, subsample = FALSE) {
  
  # Replicability
  set.seed(seed)
  
  # Create blind code
  blind <- c(LETTERS, str_c(LETTERS,LETTERS), str_c(LETTERS,LETTERS,LETTERS), str_c(LETTERS,LETTERS,LETTERS,LETTERS))
  
  # Sample without replacement
  # Size in sample function equal to number of subjects to blind
  if(subsample){
    subsample <- blind[subsample[1]:subsample[2]]
    blind_rand <- sample(subsample, 
                         replace = FALSE)
    }
  else blind_rand <- sample(blind[1:nrow(subjects)], 
                       size = nrow(subjects), 
                       replace = FALSE)
  
  # Add blind code column to subject data
  subjects_blinded <- subjects |> mutate(blind_code = blind_rand)
  
  return(subjects_blinded)
}
```

### 2.1 Blinding subjects for Iba1 DAB
**2025-12-30**

Dependency: {tidyverse}

Add *unblind* column to easily rename files with meta data post-imaging

```{r}
# Read in subject data, filter out excluded subjects, select columns of interest, and add leading 0s to subjects
subjects <- read_csv("data/subjects.csv", col_names = TRUE) |> 
  filter(exclude == "N") |> 
  select(subject, group, sex) |> 
  mutate(subject = str_pad(subject, 2, pad = "0"))

# Create tibble of blinded subjects using `blind_subjects()`
subjects_blinded <- blind_subjects(subjects = subjects, seed = 7)

# Split group column into drug and diet columns
subjects_blinded <- subjects_blinded |> 
  separate_wider_delim(group, delim = "-", names = c("drug", "diet")) |> 
  unite(unblind, c(subject, sex, drug, diet), sep = "_", remove = FALSE)

# Save blind key csv file
write_csv(subjects_blinded, "data/iba1_blind_key.csv")
```

### 2.2 Blinding subjects for FJB
**2026-01-06**

```{r}
# Read in subject data, filter out excluded subjects, select columns of interest, and add leading 0s to subjects
subjects <- read_csv("data/subjects.csv", col_names = TRUE) |> 
  filter(exclude == "N") |> 
  select(subject, group, sex) |> 
  mutate(subject = str_pad(subject, 2, pad = "0"))

# Create tibble of blinded subjects using `blind_subjects()`
subjects_blinded <- blind_subjects(subjects = subjects, seed = 10)

# Split group column into drug and diet columns
subjects_blinded <- subjects_blinded |> 
  separate_wider_delim(group, delim = "-", names = c("drug", "diet")) |> 
  unite(unblind, c(subject, sex, drug, diet), sep = "_", remove = FALSE)

# Save blind key csv file
write_csv(subjects_blinded, "data/fjb_blind_key.csv")
```

> **NOTE:** Compiled count data unblinded in "MGD56_FJB.qmd"


## 3. Unblinding images post-imaging, pre-analysis

### 3.1 Unblinding Iba1 DAB image file names
**2026-01-05**

Ensure all original image files are in one folder.

Input file name example:
- Stain_Study_BlindCode_BrainRegion_Slide-Slice_Magnification.FileType
- Iba1_MGD5_AA_DGML_1-10_20X.tif

Output file name example:
- Stain_Study_RatID_Sex_Drug_Diet_BrainRegion_Slide-Slice_Magnification.FileType
- Iba1_MGD5_53_F_PLX_CON_DGML_1-10_20X.tif

Goal: Create image copy with BlindCode replaced by RatID_Sex_Drug_Diet in file name
- Using columns *blind_code* and *unblind* of `blind_key` (e.g., .csv from `subjects_blinded`) 

Write `split_images()` function with the following arguments:
- *blind_key*, a data table containing the blind code and unblinded information
- *path_original*, a path to the original files
- *path_blind*, a path to the renamed files
Returns `renamed_images` data table with columns *image_name* (input file name), *new_name* (new file name), *path_input*, *path_output*

Write `renamer()` function, passing in `renamed_images` and using column names as arguments to copy files and give new names.


```{r}
# Function to split image name, bind unblind information, and create new image name
split_images <- function(blind_key, path_input, path_output) {
  
  # Create vector of file names to be renamed
  image_name <- dir(path = path_input,
           recursive = TRUE)

  # Create the tibble and process the data
  images <- tibble(image_name) |>  
    separate_wider_delim(image_name, delim = "_", 
                         names = c("stain", "study", "blind_code", "brain_region", "slideslice", "magnification"), 
                         cols_remove = FALSE)
  
  # Read in blind key containing blind_code and unblind columns
  unblind <- blind_key |> select(unblind, blind_code)
  
  # Bind blind_key data frame to images data frame
  renamed_images <- left_join(images, unblind, by = 'blind_code')
  
  # Consolidate various columns into new image name column
  # Remove blind_code column
  # Add input and output path columns to file
  renamed_images <- renamed_images |> 
    select(!blind_code) |> 
    unite("new_name", stain, study, unblind, brain_region, slideslice, magnification, sep = "_") |> 
    mutate(path_input = path_input,
           path_output = path_output)
  
  return(renamed_images)
}
```

```{r}
# Function to rename for blinding
# Arguments obtained from columns of `renamed_images` data frame created by `split_images()`
renamer <- function(image_name, new_name, path_input, path_output) {
  file.copy(from = str_c(path_input, image_name),
            to = str_c(path_output, new_name))
}
```


Test `split_images()` and `renamer()` functions

```{r include = FALSE}
# Set location of data
path <- "/Users/erc2866/Library/CloudStorage/Box-Box/Nixon Lab/Lab Member Data Files/Erika/MGD/MGD56/MicrogliaMorphology/"

# Specify location of images
path_in <- str_c(path, "test/")

path_out <- str_c(path, "test_unblind/")

# Load blind key
subjects_blinded <- read_csv("data/iba1_blind_key.csv")

# Create `<blind>` folder if it does not already exist
create_folder(path_out)

# Create tibble of split image names, including sex/diet/drug, using `split_images()`
renamed_images <- split_images(blind_key = subjects_blinded,
                       path_input = path_in, path_output = path_out) 


# Unblind images
pmap(renamed_images, renamer) # function applied to each portion of data frame
```

#### Run functions to unblind image names
**2026-04-12**

Notes

- There is a single LLL image, taken 2/26/26 between L and MM
- Iba1_MGD56_R_dHippo_1-10_20X, todo: rename dHippo to **DGML**


```{r}
# Set location of data
path <- "/Users/erc2866/Library/CloudStorage/Box-Box/Nixon Lab/Lab Member Data Files/Erika/MGD/MGD56/MicrogliaMorphology/"

# Specify location of images
path_in <- str_c(path, "images_blind/") # customize

path_out <- str_c(path, "images_unblind/") # customize

# Load blind key
subjects_blinded <- read_csv("data/iba1_blind_key.csv")

# Create `<blind>` folder if it does not already exist
create_folder(path_out)

# Create tibble of split image names, including sex/diet/drug, using `split_images()`
renamed_images <- split_images(blind_key = subjects_blinded,
                       path_input = path_in, path_output = path_out) 


# Unblind images
pmap(renamed_images, renamer) # function applied to each portion of data frame
```
