---
title: "Subject_assignment"
author: "Erika Carlson"
date: "2025-07-15"
format: html
editor_options: 
  chunk_output_type: console
---

------

This document contains research notes and code generated by:

> **Erika R. Carlson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> erika.carlson@utexas.edu

------

# Subject assignment

## Preliminaries

- Load {tidyverse}

```{r}
#| include: false
#| message: false
library(tidyverse)
```

## Randomize subject numbers and treatment

`assign_subject_num()` is a function that pseudo-randomly assigns a treatment 
group and number to each cage, then populates subject numbers based on the cage number.

It satisfies the following conditions:

- Separate treatment assignments per grouping variable (e.g., sex)
- Cage mates have sequential numbers

It has the following arguments: 

- `df` is a ".csv" with subject meta data, including **cage_num** column
- `seed` sets seed for reproducibility
- `grouping_var` specifies column(s) of **df** to group by (e.g., **sex**) prior to assignments. In quotes.
- `groups` is a vector specifying treatment groups in quotes
- `n_cages_per_group` specifies number of cages per group 
    - equal to half # of subjects per group e.g., 2 cages of PLX-CON females
    - can be the same for all groups (e.g., `2`) or a vector of numbers for custom assignment (e.g., `c(2, 4, 3, 4)`)

The function does the following:

- Uses `set.seed()` to define the initial state of the random number generator
- Creates **assigned_df** data frame from input `df` by:
  - Restricting `df` to unique values of **cage_num**, retaining `grouping_var` column(s)
  - Creating **cage_rand** column with random number assignment
  - Grouping by `grouping_var` column(s)
  - Creating **group** column with random group assignment per cage, number indicated by `n_cages_per_group `
  - Joining newly created columns to original `df`
  - Arranging data frame by **cage_rand** order
  - Creating **subject** column from row number, which now aligns with the new cage order so cage mates will have sequential numbers
  - Arranging data frame by `grouping_var` column(s)

The output is a data frame binding **group**, **cage_num**, **cage_rand**, and **subject** columns to `df`.

**Limitation:** Post-function modification required if subject numbers do not start with #1

- e.g., if starting with subject #33
- `assigned_subjects <- assigned_subjects %>% mutate(subject = subject + 32)`

It requires {dplyr} and {rlang} packages within {tidyverse}.

```{r create assign_subject_num function}
assign_subject_num <- function(df, seed = 1, grouping_var, groups, n_cages_per_group) {
  
  # Replicability
  set.seed(seed)
  
  # Convert character or unquoted vars to symbols for tidy evaluation
  grouping_var_sym <- syms(grouping_var)
  
  assigned_df <- df %>% 
    distinct(cage_num, !!!grouping_var_sym) %>%
    mutate(cage_rand = sample(cage_num, replace = FALSE)) %>% 
    group_by(!!!grouping_var_sym) %>%
    mutate(group = sample(rep(groups, times = n_cages_per_group))) %>% 
    left_join(df) %>% 
    arrange(cage_rand) %>% 
    rowid_to_column(var = "subject") %>% 
    arrange(!!!grouping_var_sym)
  
  return(assigned_df)
}
```

## MGD5

Read in subject meta data file.
```{r read in data file}
subjects <- read_csv("data/MGD5_unassigned_subjects.csv")
```

Use `assign_subject_num()` function to pseudo-randomly assign treatment groups and subject numbers.
```{r use assign_subject_num function}
assigned_subjects <- assign_subject_num(df = subjects, seed = 485, 
                                        grouping_var = "sex", 
                                        groups = c("PLX-EtOH", "PLX-CON", "VEH-EtOH", "VEH-CON"),
                                        n_cages_per_group = 3)

# reorder data frame
assigned_subjects <- assigned_subjects %>%
  select(subject, group, sex, cage_num, temp_subject, cage_rand)

assigned_subjects
```

Renumber certain subjects to distribute treatments among 1-24 and 25-48 (staggering)

Old_num			New_num
13	PLX-EtOH	M	35
14	PLX-EtOH	M	36
21	PLX-CON	M	31
22	PLX-CON	M	32
31	PLX-CON	F	21
32	PLX-CON	F	22
35	VEH-CON	M	13
36	VEH-CON	M	14

```{r}
assigned_subjects <- assigned_subjects %>% 
  mutate(subject = case_when(subject == 13 ~ 35,
                             subject == 14 ~ 36,
                             subject == 21 ~ 31,
                             subject == 22 ~ 32,
                             subject == 31 ~ 21,
                             subject == 32 ~ 22,
                             subject == 35 ~ 13,
                             subject == 36 ~ 14,
                             TRUE ~ subject))
```

Renumber certain *temp_subject* to distribute weights among treatment groups

old_temp	new_temp
37	29
38	30
47	31
48	32
41	27
42	28
27	37
28	38
31	47
32	48
29	41
30	42

```{r}
assigned_subjects <- assigned_subjects %>% 
  mutate(temp_subject = case_when(temp_subject == 37 ~ 29,
                                  temp_subject == 38 ~ 30,
                                  temp_subject == 47 ~ 31,
                                  temp_subject == 48 ~ 32,
                                  temp_subject == 41 ~ 27,
                                  temp_subject == 42 ~ 28,
                                  temp_subject == 27 ~ 37,
                                  temp_subject == 28 ~ 38,
                                  temp_subject == 31 ~ 47,
                                  temp_subject == 32 ~ 48,
                                  temp_subject == 29 ~ 41,
                                  temp_subject == 30 ~ 42,
                                  TRUE ~ temp_subject))
```

Save **assigned_subjects** to files.
```{r save data file}
write_csv(assigned_subjects, "data/MGD5_subjects.csv")
```


## MGD6

Repeat workflow from MGD5.
```{r}
subjects <- read_csv("data/MGD6_unassigned_subjects.csv")

assigned_subjects <- assign_subject_num(df = subjects, seed = 365, 
                                        grouping_var = "sex", 
                                        groups = c("PLX-EtOH", "PLX-CON", "VEH-EtOH", "VEH-CON"),
                                        n_cages_per_group = 2)

# reorder data frame
assigned_subjects <- assigned_subjects %>%
  select(subject, group, sex, cage_num, temp_subject, cage_rand)

# renumber to continue sequence from MGD5
assigned_subjects <- assigned_subjects %>% 
  mutate(subject = subject + 48)

assigned_subjects
```

Renumbered certain *temp_subject* to distribute weights among treatment groups, along with associated *cage_num* and *cage_rand*

old_temp	new_temp
11	01
12	02
01	11
02	12

```{r}
assigned_subjects <- assigned_subjects %>% 
  ungroup %>% 
  # First swap the temp_subject identifiers
  mutate(
    temp_subject = case_when(
      temp_subject == 1  ~ 11,
      temp_subject == 11 ~ 1,
      temp_subject == 2  ~ 12,
      temp_subject == 12 ~ 2,
      TRUE ~ temp_subject
    )
  ) %>%
  # Then swap corresponding cage info
  mutate(
    cage_num = case_when(
      temp_subject == 1  ~ cage_num[temp_subject == 11],
      temp_subject == 11 ~ cage_num[temp_subject == 1],
      temp_subject == 2  ~ cage_num[temp_subject == 12],
      temp_subject == 12 ~ cage_num[temp_subject == 2],
      TRUE ~ cage_num
      )
    ) %>% 
  mutate(
    cage_rand = case_when(
      temp_subject == 1  ~ cage_rand[temp_subject == 11],
      temp_subject == 11 ~ cage_rand[temp_subject == 1],
      temp_subject == 2  ~ cage_rand[temp_subject == 12],
      temp_subject == 12 ~ cage_rand[temp_subject == 2],
      TRUE ~ cage_rand
      )
    )
```

Save **assigned_subjects** to files.
```{r save data file}
write_csv(assigned_subjects, "data/MGD6_subjects.csv")
```

