---
title: "Subject_assignment"
author: "Erika Carlson"
date: "2025-07-15"
format: html
editor_options: 
  chunk_output_type: console
---

------

This document contains research notes and code generated by:

> **Erika R. Carlson**  
> Pharmacology & Toxicology PhD Student  
> University of Texas at Austin  
> erika.carlson@utexas.edu

------

# Subject assignment {#module-1}

## Preliminaries

- Load {tidyverse}

```{r}
#| include: false
#| message: false
library(tidyverse)
```

## Randomize subject numbers and treatment

`assign_subject_num()` is a function that pseudo-randomly assigns a treatment 
group and number to each cage, then populates subject numbers based on the cage number.

It satisfies the following conditions:

- Equal number of treatment assignments per grouping variable (e.g., sex)
- Cage mates have sequential numbers

It has the following arguments: 

- `df` is a ".csv" with subject meta data, including **cage_num** column
- `seed` sets seed for reproducibility
- `grouping_var` specifies column of **df** to group by (e.g., **sex**) prior to assignments
- `groups` is a vector specifying treatment groups
- `n_cages_per_group` specifies number of cages per group (half # of subjects), 
can be the same for all groups (e.g., 2) or a vector of numbers for custom assignment (e.g., c(2, 4, 3, 4))

The output is a data frame binding `group`, `cage_num`, `cage_rand`, and `subject` columns to `df`.

It requires {dplyr} package.

```{r create assign_subject_num function}
assign_subject_num <- function(df, seed = 1, grouping_var, groups, n_cages_per_group) {
  
  # Replicability
  set.seed(seed)
  
  assigned_df <- df %>% 
    group_by({{grouping_var}}) %>% # column of df
    distinct(cage_num) %>% 
    mutate(group = sample(rep(groups, times = n_cages_per_group)), # number of group assignments per cage
           cage_rand = sample(cage_num, replace = FALSE)) %>% 
    left_join(subjects) # joining by common columns including cage_num
  
  assigned_df <- assigned_df %>% arrange(cage_rand)
  
  assigned_df <- rowid_to_column(assigned_df, var = "subject")
  
  return(assigned_df)
}
```

Read in subject meta data file
```{r use assign_subject_num function}
subjects <- read_csv("/Users/erc2866/Library/CloudStorage/Box-Box/Nixon Lab/Binge Files/MGD/MGD5/MGD5_unassigned_subjects.csv")
```

Use `assign_subject_num()` function to pseudo-randomly assign treatment groups and subject numbers
```{r}
assigned_subjects <- assign_subject_num(df = subjects, seed = 435, grouping_var = sex, 
                                        groups = c("PLX-EtOH", "PLX-CON", "VEH-EtOH", "VEH-CON"),
                                        n_cages_per_group = 2)

assigned_subjects
```
